---
title: "FINAL_18S_Taxonomy"
author: "Nicole Pittoors"
date: "2025-09-05"
output: html_document
---
Load Libraries

```{r, load libraries}
library(tidyverse)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(vegan)
```

# Set working directory

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "Set/Path")

# Also set working directory for interactive use
setwd("Set/Path")

# Set output directory
output_dir <- "./Taxa_Output_18S"

# Create the directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Verify it was created
cat("Output directory set to:", output_dir, "\n")
cat("Full path:", normalizePath(output_dir), "\n")

```


# Import data
```{r import data}
# Load ASV proportions table, 4th root transformed, no ASVS with zero reads across all samples, ordered west to east and then by sample type, rows are SHIDs, matched metadata
load("./biotic_percs_18S_4th.Rdata")
# ASV raw count table, no ASVs with zero reads across all samples
load("./biotic_clean_18S.Rdata")
# Metadata
load("./metadata_18S.Rdata")
# Metazoan taxonomy table
metazoo_tax_tab0 <- read.delim("./taxa_metazoan_18S.tsv", row.names=1) 
dim(metazoo_tax_tab0) # 7089 x 163

# Remove any ASVs in taxa table that don't match rownames of biotic_percs_18th_4th
metazoo_tax_tab <- metazoo_tax_tab0[match(rownames(biotic_percs_18S_4th), rownames(metazoo_tax_tab0)), ]
dim(metazoo_tax_tab) #6332 x 24
# Extract only the taxonomic hierarchy columns from taxa table
taxonomic_columns <- c("phylum", "class", "order", "family", "genus", "species")
metazoo_tax_tab <- as.matrix(metazoo_tax_tab[, taxonomic_columns])

```

# Create Taxonomy Phyloseq object
```{r create taxonomy phyloseq object}
# Create phyloseq object with this basic sample data
physeq <- phyloseq(
  otu_table(biotic_percs_18S_4th, taxa_are_rows = TRUE),
  sample_data(metadata_18S),
  tax_table(metazoo_tax_tab)
)

# Check if that worked
str(physeq)

# Save
```

# Calculate ASV abundances
Calculate the maximum abundance of each ASV across all samples and then see how many ASVs have a maximum abundance greater than 10% in any sample
Ideal when looking at lower level taxa than just Phyla
```{r ASV abundances}
# Calculate ASV abundances
#ASV_Abundance <- rowSums(biotic_clean_18S)

# Calculate relative abundances by sample (column)
# Convert counts to percentages within each sample
#sample_totals <- colSums(otu_matrix)
#metazo_tpercs <- t(apply(otu_matrix, 1, function(x) x*100/sample_totals))

# Step 3: Find the maximum percentage for each ASV across all samples
#asv_max_percent <- apply(metazo_tpercs, 1, max)

# Step 4: Count how many ASVs have a maximum abundance > 10%
#sum(asv_max_percent > 10) # 49

# You can also see which ASVs these are
#dominant_asvs <- names(asv_max_percent[asv_max_percent > 10])
#head(dominant_asvs) 

# And get their taxonomic information
#tax_info_dominant <- tax_table(physeq)[dominant_asvs, ]
#head(tax_info_dominant)

#-------- Above code but working with already imported data sets
#Convert proportions to percentages 
biotic_percs_18S_4th <- biotic_percs_18S_4th *100

# Find max percentages per ASV across samples
asv_max_percent <- apply(biotic_percs_18S_4th, 1, max)

# Count how many ASVs have a maximum abundance > 10%
sum(asv_max_percent > 10)

# You can also see which ASVs these are
dominant_asvs <- names(asv_max_percent[asv_max_percent > 10])
head(dominant_asvs) 

# And get their taxonomic information
tax_info_dominant <- tax_table(physeq)[dominant_asvs, ]
head(tax_info_dominant)
```
==========================================================
--------------RELATIVE ABUNDANCE PLOTS-------------------
==========================================================

# Custom sample order based on turbidity values
```{r custom sample order}
custom_order_turb <- c(
   # MCG
  "16_MCG_100", "17_MCG_100", "18_MCG_100",
  "16_MCG_500", "17_MCG_500", "18_MCG_500",
  "16_MCG_SESSILE", "17b_MCG_SESSILE", "18_MCG_SESSILE",
  
  # EFGshal
  "31_EFGshal_100", "32_EFGshal_100", "33_EFGshal_100",
  "31_EFGshal_500", "32_EFGshal_500", "33_EFGshal_500",
  "31_EFGshal_SESSILE", "32_EFGshal_SESSILE", "33_EFGshal_SESSILE",
 
   # ALDshal
  "13_ALDshal_100", "14_ALDshal_100", "15_ALDshal_100",
  "13_ALDshal_500", "14_ALDshal_500", "15_ALDshal_500",
  "13_ALDshal_SESSILE", "14_ALDshal_SESSILE", "15_ALDshal_SESSILE",
 
   # BRIshal
  "25_BRIshal_100", "26_BRIshal_100", "26c_BRIshal_100", "27_BRIshal_100",
  "25_BRIshal_500", "26_BRIshal_500", "26b_BRIshal_500", "27_BRIshal_500",
  "25_BRIshal_SESSILE", "26_BRIshal_SESSILE", "27_BRIshal_SESSILE",
  
  # DIAback
  "01_DIAback_100", "02_DIAback_100", "03_DIAback_100",
  "01_DIAback_500", "02_DIAback_500", "03_DIAback_500",
  "01_DIAback_SESSILE", "02_DIAback_SESSILE", "03_DIAback_SESSILE",
  
   # EFGdeep
  "34_EFGdeep_100", "35_EFGdeep_100", "36_EFGdeep_100",
  "34_EFGdeep_500", "35_EFGdeep_500", "36_EFGdeep_500",
  "34_EFGdeep_SESSILE", "35_EFGdeep_SESSILE", "36_EFGdeep_SESSILE",
  
  # DIAcoral
  "04_DIAcoral_100", "05_DIAcoral_100", "06_DIAcoral_100",
  "04_DIAcoral_500", "05_DIAcoral_500", "06_DIAcoral_500",
  "04_DIAcoral_SESSILE", "05_DIAcoral_SESSILE", "06_DIAcoral_SESSILE",
  
  # BRIback
  "22_BRIback_100", "23_BRIback_100", "24_BRIback_100",
  "22_BRIback_500", "23_BRIback_500", "24_BRIback_500",
  "22_BRIback_SESSILE", "23_BRIback_SESSILE", "24_BRIback_SESSILE",
  
  # BRIcoral
  "19_BRIcoral_100", "20_BRIcoral_100", "20b_BRIcoral_100", "21_BRIcoral_100",
  "19_BRIcoral_500", "20_BRIcoral_500", "21_BRIcoral_500",
  "19_BRIcoral_SESSILE", "20_BRIcoral_SESSILE", "21_BRIcoral_SESSILE",
  
   # ALDcoral
  "07_ALDcoral_100", "08_ALDcoral_100", "09_ALDcoral_100",
  "07_ALDcoral_500", "08_ALDcoral_500", "09_ALDcoral_500",
  "07_ALDcoral_SESSILE", "08_ALDcoral_SESSILE", "09_ALDcoral_SESSILE",
  
  # STE
  "28_STE_100", "29_STE_100", "29b_STE_100", "30_STE_100", "30b_STE_100",
  "28_STE_500", "29_STE_500", "30_STE_500", "30b_STE_500",
  "28b_STE_SESSILE", "29b_STE_SESSILE", "30_STE_SESSILE",
  
  # ALDback
  "10_ALDback_100", "11_ALDback_100", "12_ALDback_100",
  "10_ALDback_500", "11_ALDback_500", "12_ALDback_500",
  "10_ALDback_SESSILE", "11_ALDback_SESSILE", "12_ALDback_SESSILE"
)
```


# Create color palette that matches pre-taxa figure
```{r color palette}
c20 <- c("#60714E","#015951","#3B8C6E","#B0DBBB","#5BBFB0","#9DC9D3","#408BA8","#0D3D59","#594A9E","#A8B1D6","#812A5A","#F69B9C","#F16363","#C74238","#FEE14A","#FEBC2C","#F68C30","#F05522","#FDCA9D","grey40")


# Create Phyla mapped palette USE THIS ONE
phylum_color_map <- c("Annelida" = "#60714E",
                      "Chaetognatha" = "#015951",
                      "Arthropoda" = "#3B8C6E",
                      "Sipuncula" = "#B0DBBB",
                      "Bryozoa" = "#5BBFB0",
                      "Brachiopoda" = "#9DC9D3",
                      "Chordata" = "#408BA8",
                      "Rotifera" = "#0D3D59",
                      "Cnidaria" = "#594A9E",
                      "Echinodermata" = "#A8B1D6",
                      "Entoprocta" = "#812A5A",
                      "Mollusca" = "#F69B9C",
                      "Rhodophyta" = "#F16363",
                      "Nematoda" = "#C74238",
                      "Nemertea" = "#FEE14A",
                      "Platyhelminthes" = "#FEBC2C",
                      "Phoronida" = "#C47B51",
                      "Porifera" = "#F68C30",
                      "Ctenophora" = "#F05522",
                      "Gastrotricha" = "#FDCA9D",
                      "Acanthocephala" = "#104820",
                      "Xenacoelomorpha" = "#C47B51",
                      "Unknown" = "grey40")
```

# Create outut directory
```{r output directory}
dir.create("./Taxa_Output_18S", showWarnings = FALSE)

```


# Relative Abundance Plot by SAMPLE
```{r rel. abund. plot by SAMPLE}
# Start with the phyloseq object
otu_matrix <- as(otu_table(physeq), "matrix")
tax_matrix <- as(tax_table(physeq), "matrix")
sample_data_df <- as.data.frame(sample_data(physeq))

# First, create a data frame with the sample metadata we need
sample_metadata <- data.frame(
  Sample = rownames(sample_data_df),
  sample_id_r = as.character(sample_data_df$sample_id_r),  # Convert to character to avoid list issues
  fraction = gsub(".*_(\\d+|SESSILE)$", "\\1", as.character(sample_data_df$sample_id_r)),
  sitelocality = as.character(sample_data_df$sitelocality),
  stringsAsFactors = FALSE
)

# Calculate phylum-level abundances directly
# First, group ASVs by phylum
phylum_groups <- split(rownames(tax_matrix), tax_matrix[, "phylum"])

# For each phylum, sum the abundances of its ASVs for each sample
phylum_abundances <- lapply(names(phylum_groups), function(phylum) {
  asvs <- phylum_groups[[phylum]]
  if (length(asvs) == 1) {
    # If only one ASV in phylum, just get its abundance
    abundance <- otu_matrix[asvs, , drop = FALSE]
  } else {
    # Otherwise sum the abundances of all ASVs in the phylum
    abundance <- colSums(otu_matrix[asvs, , drop = FALSE])
  }
  data.frame(
    Sample = colnames(otu_matrix),
    phylum = phylum,
    Abundance = as.numeric(abundance),
    stringsAsFactors = FALSE
  )
})

# Combine all phylum abundances
phylum_abundance_df <- do.call(rbind, phylum_abundances)

# Merge with sample metadata
phylum_abundance_df <- merge(phylum_abundance_df, sample_metadata, by = "Sample")

# Calculate total abundance by phylum
phylum_totals <- aggregate(Abundance ~ phylum, data = phylum_abundance_df, sum)
phylum_totals <- phylum_totals[order(phylum_totals$Abundance, decreasing = TRUE), ]

# Get top 20 phyla
top_20_phyla <- phylum_totals$phylum[1:20]


# Filter to only include top 20 phyla
phylum_abundance_df_filtered <- phylum_abundance_df[phylum_abundance_df$phylum %in% top_20_phyla, ]

# Calculate relative abundance
library(dplyr)
relative_data <- phylum_abundance_df_filtered %>%
  group_by(Sample) %>%
  mutate(Relative_Abundance = Abundance / sum(Abundance) * 100) %>%
  ungroup()

# Order samples by custom order
relative_data$sample_id_r <- factor(relative_data$sample_id_r, levels = custom_order_turb)

# Create variable to label each sample by sitelocality value
# First create a named vector mapping sample_id_r to site
site_labels <- setNames(metadata_ARMS$sitelocality, metadata_ARMS$sample_id_r)
# Remove duplicates if any
site_labels <- site_labels[!duplicated(names(site_labels))]

# Create the plot
p <- ggplot(relative_data, aes(x = sample_id_r, y = Relative_Abundance, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack", 
           color = "black", size = 0.2, width = 0.8) +
  facet_grid(~ fraction, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = phylum_color_map) +
  scale_x_discrete(labels = site_labels) +  # This maps sample_id_r to site labels
  labs(y = "Relative abundance (%)", x = "Site", title = "Relative Abundance of Top 20 Phyla") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 11, color = "black"),
    axis.text.y = element_text(size = 20, color = "black"),
    axis.title = element_text(size = 17, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 17, face = "bold", margin = margin(t = 0, b = 5)),
    legend.title = element_text(size = 17, face = "bold", color = "black"),
    legend.text = element_text(size = 16, color = "black"),
    legend.position = "right",
    panel.grid = element_blank(),
    panel.spacing = unit(1, "cm")
  )

# Save the plot
ggsave("./18S_Top20_Phyla_RelativeAbundance_bySample.pdf", p, width = 18, height = 8)
```

# Relative Abundance Plot - Samples aggregated by SITELOCALITY
This code includes some unclassified samples (maybe phylum is missing?)
```{r Relative Abundance Plot - Samples aggregated by SITELOCALITY}
# Plot top 20 phyla by sitelocality, ordered by turbidity

# Aggregate the data by site locality and fraction
site_aggregated <- relative_data %>%
  group_by(sitelocality, fraction, phylum) %>%
  summarise(Relative_Abundance = mean(Relative_Abundance, na.rm = TRUE)) %>%
  ungroup()

# Create a sensible ordering for site localities
siteloc_order_turb <- c("McGrail_deep","EFGB_shallow","Alderdice_shallow","Bright_shallow",
                        "Diaphus_deep2", "EFGB_deep", "Diaphus_deep1","Bright_deep2",
                        "Bright_deep1", "Alderdice_deep1", "Stetson_shallow","Alderdice_deep2")

# Set factor order
site_aggregated$sitelocality <- factor(site_aggregated$sitelocality, levels = siteloc_order_turb)

# Create the site-level plot
p_site <- ggplot(site_aggregated, aes(x = sitelocality, y = Relative_Abundance, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack",
           color = "black", size = 0.4, width = 0.85) +
  facet_grid(~ fraction, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = phylum_color_map) +
  labs(y = "Relative abundance (%)", x = "Site Locality", title = "Relative Abundance of Top 20 Phyla by Site- 18S") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12, color = "black"),
    axis.text.y = element_text(size = 16, color = "black"),
    axis.title = element_text(size = 17, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 17, face = "bold", margin = margin(t = 0, b = 5)),
    legend.title = element_text(size = 17, face = "bold", color = "black"),
    legend.text = element_text(size = 16, color = "black"),
    legend.position = "right",
    panel.grid = element_blank(),
    panel.spacing = unit(0.5, "cm")
  )

# Save the plot
ggsave("./18S_Top20_Phyla_RelativeAbundance_BySite.pdf", p_site, width = 14, height = 10)
```

Plot top 20 phyla by sitelocality, ordered by turbidity
```{r Plot top 20 phyla by sitelocality, ordered by turbidity}
# Plot top 20 phyla by sitelocality, ordered by turbidity

# Step 1: Check initial data
cat("Initial phyla in data:", 
    length(unique(phylum_abundance_df$phylum)), "\n")
print(unique(phylum_abundance_df$phylum))

# Step 2: Filter to only include top 20 phyla
phylum_abundance_df_filtered <- phylum_abundance_df[phylum_abundance_df$phylum %in% top_20_phyla, ]
cat("\nAfter filtering to top 20:", 
    length(unique(phylum_abundance_df_filtered$phylum)), "\n")

# Step 3: Remove unclassified
phylum_abundance_df_filtered <- phylum_abundance_df_filtered %>%
  filter(!is.na(phylum),
         phylum != "",
         phylum != "Unclassified",
         phylum != "unclassified")

cat("After removing unclassified:", 
    length(unique(phylum_abundance_df_filtered$phylum)), "\n")
print(sort(unique(phylum_abundance_df_filtered$phylum)))

# Step 4: Calculate relative abundance
relative_data <- phylum_abundance_df_filtered %>%
  group_by(Sample) %>%
  mutate(Relative_Abundance = Abundance / sum(Abundance) * 100) %>%
  ungroup()

# Continue with rest of code...
```

# See top 10 phyla in each fraction
```{r top 10 phyla in each fraction}
# Check phylum abundance by fraction
phylum_by_fraction <- aggregate(Relative_Abundance ~ phylum + fraction, 
                                data = relative_data, 
                                FUN = mean)

# View top phyla by fraction
for(frac in unique(phylum_by_fraction$fraction)) {
  cat("\nTop phyla in fraction", frac, ":\n")
  print(phylum_by_fraction[phylum_by_fraction$fraction == frac, ] %>%
          arrange(desc(Relative_Abundance)) %>%
          head(10))
}
```

===============================================================================
----------- Relative Abundance Table including UNCLASSIFIED READS--------------
===============================================================================
Creating a new ASV count table and phyloseq object to plot relative abundances including unclassified reads

# Load missing ASVs, original AVV count table, and original taxonomy table
```{r rel abund plot with unclassified reads}
# Load missing ASVs (those with no taxonomy hits)
missing_asvs_raw <- read.delim("./missing_asvs.tsv", header = FALSE, stringsAsFactors = FALSE)
cat("Missing ASVs file dimensions:", dim(missing_asvs_raw), "\n")
cat("First few entries:\n")
print(head(missing_asvs_raw))

# Extract just the ASV IDs (should be in first column)
missing_asv_ids <- missing_asvs_raw$V1
cat("\nTotal missing ASVs:", length(missing_asv_ids), "\n")

# Load original count table
original_counts <- read.csv('/Users/nicolepittoors/Documents/Documents - biscaya/Bioinformatics/ARMS_18S_LULU/18S_lulu_ASVtab.csv', 
                            row.names = 1)
cat("\nOriginal count table dimensions:", dim(original_counts), "\n")
cat("First few row names:", head(rownames(original_counts)), "\n")
cat("First few column names:", head(colnames(original_counts)), "\n")

# Load the COMPLETE taxonomy file (metazoan + non-metazoan)
complete_taxonomy <- read.delim('./final_taxonomy_results.tsv',
                                row.names = 1)

cat("=== Complete Taxonomy File ===\n")
cat("Dimensions:", dim(complete_taxonomy), "\n")
cat("Column names:", colnames(complete_taxonomy), "\n\n")

# Check kingdoms present
if("kingdom" %in% colnames(complete_taxonomy)) {
  cat("Kingdoms present:\n")
  print(table(complete_taxonomy$kingdom))
  cat("\n")
}

# Check phyla
if("phylum" %in% colnames(complete_taxonomy)) {
  cat("Top 20 phyla:\n")
  print(head(sort(table(complete_taxonomy$phylum), decreasing = TRUE), 20))
  cat("\n")
}
```

# Check which missing ASVs are in filtered data
```{r find missing ASVs}
cat("=== ASV Accounting ===\n")
cat("Original count table:", nrow(original_counts), "ASVs\n")
cat("Complete taxonomy file:", nrow(complete_taxonomy), "ASVs\n")
cat("Missing ASVs (no hits):", length(missing_asv_ids), "ASVs\n")

# Calculate non-metazoan
asvs_in_complete_tax <- rownames(complete_taxonomy)
non_metazoan_count <- length(setdiff(asvs_in_complete_tax, rownames(metazoo_tax_tab)))

cat("\nBreakdown:\n")
cat("  Metazoan/Rhodophyta:", nrow(metazoo_tax_tab), "ASVs\n")
cat("  Non-metazoan:", non_metazoan_count, "ASVs\n")
cat("  No taxonomy hits:", length(missing_asv_ids), "ASVs\n")
cat("  Total accounted:", nrow(metazoo_tax_tab) + non_metazoan_count + length(missing_asv_ids), "\n")

# The key question: Which missing ASVs are in ARMS samples?
cat("\n=== Missing ASVs in ARMS Samples ===\n")

# Extract SH IDs from original column names
sh_pattern <- "SH[0-9]+"
original_sh_ids <- regmatches(colnames(original_counts), 
                              regexpr(sh_pattern, colnames(original_counts)))

# Find which columns match your ARMS samples
arms_sample_ids <- colnames(biotic_clean_18S)
arms_cols_in_original <- which(original_sh_ids %in% arms_sample_ids)

cat("Columns in original matching ARMS samples:", length(arms_cols_in_original), "\n")

# Subset to ARMS samples only
original_arms_only <- original_counts[, arms_cols_in_original]
cat("Original table subset to ARMS:", dim(original_arms_only), "\n")

# Check missing ASVs in ARMS samples
missing_in_arms <- intersect(missing_asv_ids, rownames(original_arms_only))
cat("\nMissing ASVs present in original ARMS samples:", length(missing_in_arms), "\n")

if(length(missing_in_arms) > 0) {
  # Calculate their abundances
  missing_arms_abundances <- rowSums(original_arms_only[missing_in_arms, ])
  
  cat("Abundance summary for missing ASVs in ARMS:\n")
  print(summary(missing_arms_abundances))
  
  # How many have non-zero counts?
  missing_with_counts <- sum(missing_arms_abundances > 0)
  cat("\nMissing ASVs with >0 counts in ARMS:", missing_with_counts, "\n")
  cat("Missing ASVs with 0 counts in ARMS:", sum(missing_arms_abundances == 0), "\n")
  
  # These are your "Unclassified" candidates!
  if(missing_with_counts > 0) {
    unclassified_asvs <- names(missing_arms_abundances[missing_arms_abundances > 0])
    cat("\n✓ Found", length(unclassified_asvs), "unclassified ASVs with counts in ARMS samples!\n")
    
    # What percentage of total reads?
    total_reads_all <- sum(original_arms_only)
    unclassified_reads <- sum(missing_arms_abundances[missing_arms_abundances > 0])
    cat("Unclassified reads:", unclassified_reads, 
        "(", round(unclassified_reads/total_reads_all*100, 2), "% of total)\n")
  }
}
```

```{r debugging}
cat("Debugging column name mismatch:\n\n")

# Check what we have
cat("Sample of original_arms_only column names:\n")
print(head(colnames(original_arms_only)))

cat("\nSample of biotic_clean_18S column names:\n")
print(head(colnames(biotic_clean_18S)))

cat("\nSample of original_sh_ids (extracted SH IDs):\n")
print(head(original_sh_ids[arms_cols_in_original]))

cat("\nSample of arms_sample_ids:\n")
print(head(arms_sample_ids))

# The issue is likely the matching - let's fix it
# Extract SH IDs more carefully
original_arms_sh <- original_sh_ids[arms_cols_in_original]

cat("\nChecking if SH IDs are in the same order:\n")
cat("First few from original:", head(original_arms_sh), "\n")
cat("First few from biotic_clean:", head(arms_sample_ids), "\n")

# Try direct assignment if they're in the same order
if(length(original_arms_sh) == length(arms_sample_ids)) {
  cat("\nSame length - checking if same samples in same order...\n")
  cat("Proportion matching:", mean(original_arms_sh == arms_sample_ids), "\n")
  
  if(mean(original_arms_sh == arms_sample_ids) > 0.9) {
    # They're in the same order, just assign directly
    cat("✓ Samples are in same order, assigning directly\n")
    colnames(original_arms_only) <- colnames(biotic_clean_18S)
    colnames(unclassified_counts) <- colnames(biotic_clean_18S)
  } else {
    # They're not in the same order, need to reorder
    cat("✗ Samples are NOT in same order, reordering...\n")
    
    # Create a mapping
    mapping <- data.frame(
      original_col = colnames(original_arms_only),
      sh_id = original_arms_sh
    )
    
    # Match to biotic_clean order
    match_indices <- match(colnames(biotic_clean_18S), mapping$sh_id)
    
    if(any(is.na(match_indices))) {
      cat("ERROR: Some samples in biotic_clean_18S not found in original_arms_only\n")
      cat("Missing samples:", sum(is.na(match_indices)), "\n")
    } else {
      # Reorder columns
      original_arms_only <- original_arms_only[, match_indices]
      unclassified_counts <- unclassified_counts[, match_indices]
      
      # Now assign column names
      colnames(original_arms_only) <- colnames(biotic_clean_18S)
      colnames(unclassified_counts) <- colnames(biotic_clean_18S)
      
      cat("✓ Reordered and column names assigned\n")
    }
  }
} else {
  cat("ERROR: Different number of samples!\n")
  cat("Original ARMS:", length(original_arms_sh), "\n")
  cat("Biotic clean:", length(arms_sample_ids), "\n")
}

# Verify the fix
cat("\nVerification:\n")
cat("Column names now match?", identical(colnames(unclassified_counts), colnames(biotic_clean_18S)), "\n")

if(identical(colnames(unclassified_counts), colnames(biotic_clean_18S))) {
  cat("✓ SUCCESS - proceeding with combining count tables\n\n")
  
  # Now continue with the rest
  cat("Step 2: Combine classified + unclassified counts\n")
  biotic_clean_18S_w_unknown <- rbind(biotic_clean_18S, unclassified_counts)
  cat("  Original biotic_clean_18S:", dim(biotic_clean_18S), "\n")
  cat("  With unclassified added:", dim(biotic_clean_18S_w_unknown), "\n\n")
  
  # Step 3: Create taxonomy table for unclassified
  cat("Step 3: Create taxonomy for unclassified ASVs\n")
  unclassified_tax <- matrix(
    "Unclassified",
    nrow = length(unclassified_asvs),
    ncol = 6,
    dimnames = list(unclassified_asvs, c("phylum", "class", "order", "family", "genus", "species"))
  )
  
  cat("  Unclassified taxonomy table:", dim(unclassified_tax), "\n\n")
  
  # Step 4: Combine taxonomy tables
  cat("Step 4: Combine taxonomy tables\n")
  metazoo_tax_tab_w_unknown <- rbind(metazoo_tax_tab, unclassified_tax)
  cat("  Original metazoo_tax_tab:", dim(metazoo_tax_tab), "\n")
  cat("  With unclassified added:", dim(metazoo_tax_tab_w_unknown), "\n\n")
  
  # Step 5: Verify alignment
  cat("Step 5: Verify alignment\n")
  cat("  Row names match?", identical(rownames(biotic_clean_18S_w_unknown), 
                                      rownames(metazoo_tax_tab_w_unknown)), "\n")
  
  # If not, fix it
  if(!identical(rownames(biotic_clean_18S_w_unknown), rownames(metazoo_tax_tab_w_unknown))) {
    cat("  Reordering to match...\n")
    metazoo_tax_tab_w_unknown <- metazoo_tax_tab_w_unknown[rownames(biotic_clean_18S_w_unknown), ]
    cat("  After reordering, match?", identical(rownames(biotic_clean_18S_w_unknown), 
                                                 rownames(metazoo_tax_tab_w_unknown)), "\n")
  }
  
  
  # Step 6: Create new phyloseq object
  cat("Step 6: Create phyloseq object\n")
  physeq18S_w_unknown <- phyloseq(
    otu_table(biotic_clean_18S_w_unknown, taxa_are_rows = TRUE),
    sample_data(metadata_18S),
    tax_table(metazoo_tax_tab_w_unknown)
  )
  
  print(physeq18S_w_unknown)
  
  # Step 7: Summary statistics
  cat("\n=== Summary ===\n")
  phylum_table <- table(tax_table(physeq18S_w_unknown)[, "phylum"])
  cat("\nPhylum distribution (top 15):\n")
  print(head(sort(phylum_table, decreasing = TRUE), 15))
  
  unclassified_count <- sum(tax_table(physeq18S_w_unknown)[, "phylum"] == "Unclassified")
  total_asvs <- ntaxa(physeq18S_w_unknown)
  
  cat("\nASV counts:\n")
  cat("  Classified (Metazoa/Rhodophyta):", nrow(metazoo_tax_tab), "\n")
  cat("  Unclassified:", unclassified_count, "\n")
  cat("  Total:", total_asvs, "\n")
  cat("  Percentage unclassified:", round(unclassified_count/total_asvs*100, 2), "%\n")
  
  # Check read abundance
  cat("\nRead abundance:\n")
  total_reads <- sum(otu_table(physeq18S_w_unknown))
  unclassified_reads_check <- sum(otu_table(physeq18S_w_unknown)[rownames(unclassified_tax), ])
  cat("  Total reads:", total_reads, "\n")
  cat("  Unclassified reads:", unclassified_reads_check, "\n")
  cat("  Percentage unclassified reads:", round(unclassified_reads_check/total_reads*100, 2), "%\n")
  
} else {
  cat("✗ FAILED - need to investigate further\n")
}
```

# Create relative abundance plots
```{r create plot w/ unclassified reads}
# Use the same custom order and color palette
custom_order_turb <- c(
  # MCG
  "16_MCG_100", "17_MCG_100", "18_MCG_100",
  "16_MCG_500", "17_MCG_500", "18_MCG_500",
  "16_MCG_SESSILE", "17b_MCG_SESSILE", "18_MCG_SESSILE",
  
  # EFGshal
  "31_EFGshal_100", "32_EFGshal_100", "33_EFGshal_100",
  "31_EFGshal_500", "32_EFGshal_500", "33_EFGshal_500",
  "31_EFGshal_SESSILE", "32_EFGshal_SESSILE", "33_EFGshal_SESSILE",
  
  # ALDshal
  "13_ALDshal_100", "14_ALDshal_100", "15_ALDshal_100",
  "13_ALDshal_500", "14_ALDshal_500", "15_ALDshal_500",
  "13_ALDshal_SESSILE", "14_ALDshal_SESSILE", "15_ALDshal_SESSILE",
  
  # BRIshal
  "25_BRIshal_100", "26_BRIshal_100", "26c_BRIshal_100", "27_BRIshal_100",
  "25_BRIshal_500", "26_BRIshal_500", "26b_BRIshal_500", "27_BRIshal_500",
  "25_BRIshal_SESSILE", "26_BRIshal_SESSILE", "27_BRIshal_SESSILE",
  
  # DIAback
  "01_DIAback_100", "02_DIAback_100", "03_DIAback_100",
  "01_DIAback_500", "02_DIAback_500", "03_DIAback_500",
  "01_DIAback_SESSILE", "02_DIAback_SESSILE", "03_DIAback_SESSILE",
  
  # EFGdeep
  "34_EFGdeep_100", "35_EFGdeep_100", "36_EFGdeep_100",
  "34_EFGdeep_500", "35_EFGdeep_500", "36_EFGdeep_500",
  "34_EFGdeep_SESSILE", "35_EFGdeep_SESSILE", "36_EFGdeep_SESSILE",
  
  # DIAcoral
  "04_DIAcoral_100", "05_DIAcoral_100", "06_DIAcoral_100",
  "04_DIAcoral_500", "05_DIAcoral_500", "06_DIAcoral_500",
  "04_DIAcoral_SESSILE", "05_DIAcoral_SESSILE", "06_DIAcoral_SESSILE",
  
  # BRIback
  "22_BRIback_100", "23_BRIback_100", "24_BRIback_100",
  "22_BRIback_500", "23_BRIback_500", "24_BRIback_500",
  "22_BRIback_SESSILE", "23_BRIback_SESSILE", "24_BRIback_SESSILE",
  
  # BRIcoral
  "19_BRIcoral_100", "20_BRIcoral_100", "20b_BRIcoral_100", "21_BRIcoral_100",
  "19_BRIcoral_500", "20_BRIcoral_500", "21_BRIcoral_500",
  "19_BRIcoral_SESSILE", "20_BRIcoral_SESSILE", "21_BRIcoral_SESSILE",
  
  # ALDcoral
  "07_ALDcoral_100", "08_ALDcoral_100", "09_ALDcoral_100",
  "07_ALDcoral_500", "08_ALDcoral_500", "09_ALDcoral_500",
  "07_ALDcoral_SESSILE", "08_ALDcoral_SESSILE", "09_ALDcoral_SESSILE",
  
  # STE
  "28_STE_100", "29_STE_100", "29b_STE_100", "30_STE_100", "30b_STE_100",
  "28_STE_500", "29_STE_500", "30_STE_500", "30b_STE_500",
  "28b_STE_SESSILE", "29b_STE_SESSILE", "30_STE_SESSILE",
  
  # ALDback
  "10_ALDback_100", "11_ALDback_100", "12_ALDback_100",
  "10_ALDback_500", "11_ALDback_500", "12_ALDback_500",
  "10_ALDback_SESSILE", "11_ALDback_SESSILE", "12_ALDback_SESSILE"
)

# Color palette with Unclassified
phylum_color_map_w_unknown <- c(
  "Annelida" = "#60714E",
  "Arthropoda" = "#3B8C6E",
  "Sipuncula" = "#B0DBBB",
  "Bryozoa" = "#5BBFB0",
  "Brachiopoda" = "#9DC9D3",
  "Chordata" = "#408BA8",
  "Rotifera" = "#0D3D59",
  "Cnidaria" = "#594A9E",
  "Echinodermata" = "#A8B1D6",
  "Entoprocta" = "#812A5A",
  "Mollusca" = "#F69B9C",
  "Rhodophyta" = "#F16363",
  "Nematoda" = "#C74238",
  "Nemertea" = "#FEE14A",
  "Platyhelminthes" = "#FEBC2C",
  "Porifera" = "#F68C30",
  "Gastrotricha" = "#FDCA9D",
  "Xenacoelomorpha" = "#C47B51",  
  "Acanthocephala" = "#104820",   
  "Unclassified" = "grey70"
)

# Calculate phylum-level abundances from physeq18S_w_unknown
otu_matrix <- as(otu_table(physeq18S_w_unknown), "matrix")
tax_matrix <- as(tax_table(physeq18S_w_unknown), "matrix")
sample_data_df <- as.data.frame(sample_data(physeq18S_w_unknown))

# Create sample metadata
sample_metadata <- data.frame(
  Sample = rownames(sample_data_df),
  sample_id_r = as.character(sample_data_df$sample_id_r),
  fraction = gsub(".*_(\\d+|SESSILE)$", "\\1", as.character(sample_data_df$sample_id_r)),
  sitelocality = as.character(sample_data_df$sitelocality),
  stringsAsFactors = FALSE
)

# Group ASVs by phylum
phylum_groups <- split(rownames(tax_matrix), tax_matrix[, "phylum"])

# Calculate phylum abundances
phylum_abundances <- lapply(names(phylum_groups), function(phylum) {
  asvs <- phylum_groups[[phylum]]
  if (length(asvs) == 1) {
    abundance <- otu_matrix[asvs, , drop = FALSE]
  } else {
    abundance <- colSums(otu_matrix[asvs, , drop = FALSE])
  }
  data.frame(
    Sample = colnames(otu_matrix),
    phylum = phylum,
    Abundance = as.numeric(abundance),
    stringsAsFactors = FALSE
  )
})

phylum_abundance_df <- do.call(rbind, phylum_abundances)
phylum_abundance_df <- merge(phylum_abundance_df, sample_metadata, by = "Sample")

# Get top 20 phyla (which now includes Unclassified)
phylum_totals <- aggregate(Abundance ~ phylum, data = phylum_abundance_df, sum)
phylum_totals <- phylum_totals[order(phylum_totals$Abundance, decreasing = TRUE), ]
top_20_phyla <- phylum_totals$phylum[1:20]

cat("Top 20 phyla (including Unclassified):\n")
print(phylum_totals[1:20, ])

# Filter to top 20
phylum_abundance_df_filtered <- phylum_abundance_df[phylum_abundance_df$phylum %in% top_20_phyla, ]

# Calculate relative abundance
library(dplyr)
relative_data <- phylum_abundance_df_filtered %>%
  group_by(Sample) %>%
  mutate(Relative_Abundance = Abundance / sum(Abundance) * 100) %>%
  ungroup()

# Order samples
relative_data$sample_id_r <- factor(relative_data$sample_id_r, levels = custom_order_turb)

# Create site labels
site_labels <- setNames(metadata_18S$sitelocality, metadata_18S$sample_id_r)
site_labels <- site_labels[!duplicated(names(site_labels))]

# Create the plot
p_with_unknown <- ggplot(relative_data, aes(x = sample_id_r, y = Relative_Abundance, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack", 
           color = "black", size = 0.2, width = 0.8) +
  facet_grid(~ fraction, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = phylum_color_map_w_unknown) +
  scale_x_discrete(labels = site_labels) +
  labs(y = "Relative abundance (%)", x = "Site", 
       title = "Relative Abundance of Top 20 Phyla (Including Unclassified) - 18S") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 11, color = "black"),
    axis.text.y = element_text(size = 20, color = "black"),
    axis.title = element_text(size = 17, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 17, face = "bold", margin = margin(t = 0, b = 5)),
    legend.title = element_text(size = 17, face = "bold", color = "black"),
    legend.text = element_text(size = 16, color = "black"),
    legend.position = "right",
    panel.grid = element_blank(),
    panel.spacing = unit(1, "cm")
  )

# Save the plot
ggsave(file.path(output_dir, "18S_Top20_Phyla_RelativeAbundance_bySample_WITH_UNCLASSIFIED.pdf"), 
       p_with_unknown, width = 18, height = 8)

cat("\n✓ Plot saved to:", file.path(output_dir, "18S_Top20_Phyla_RelativeAbundance_bySample_WITH_UNCLASSIFIED.pdf"), "\n")

# Display the plot
print(p_with_unknown)
```


# Checkinging into 100um fraction plotting as unknown
```{r}
cat("=== Debugging: Which phyla are missing colors? ===\n\n")

# Get all phyla actually in your top 20 plot data
phyla_in_plot <- unique(relative_data$phylum)
cat("Phyla in the plot data (top 20):\n")
print(sort(phyla_in_plot))

# Get phyla in your color palette
phyla_in_palette <- names(phylum_color_map_w_unknown)
cat("\n\nPhyla in color palette:\n")
print(sort(phyla_in_palette))

# Find which phyla are in plot but NOT in palette
missing_colors <- setdiff(phyla_in_plot, phyla_in_palette)
cat("\n\nPhyla in plot but MISSING from color palette:\n")
print(missing_colors)

# Find which phyla are in palette but NOT in plot
extra_colors <- setdiff(phyla_in_palette, phyla_in_plot)
cat("\n\nPhyla in palette but NOT in plot (these are OK):\n")
print(extra_colors)

# Check what ggplot is assigning to missing phyla
if(length(missing_colors) > 0) {
  cat("\n\n⚠️ PROBLEM: These phyla don't have assigned colors!\n")
  cat("ggplot will assign default colors to them.\n")
  cat("This is why you're seeing grey40 (ggplot's default grey)\n\n")
  
  # Show abundance of missing phyla
  for(p in missing_colors) {
    total_abund <- sum(relative_data$Relative_Abundance[relative_data$phylum == p])
    cat(p, "- Total relative abundance:", round(total_abund, 2), "%\n")
  }
}
```

# Relative Abundance Plot - Samples aggregated by SITELOCALITY (with Unclassified Reads)
```{r}
# Aggregate the data by site locality and fraction
site_aggregated <- relative_data %>%
  group_by(sitelocality, fraction, phylum) %>%
  summarise(Relative_Abundance = mean(Relative_Abundance, na.rm = TRUE)) %>%
  ungroup()

# Create a sensible ordering for site localities (by turbidity)
siteloc_order_turb <- c("McGrail_deep", "EFGB_shallow", "Alderdice_shallow", "Bright_shallow",
                        "Diaphus_deep2", "EFGB_deep", "Diaphus_deep1", "Bright_deep2",
                        "Bright_deep1", "Alderdice_deep1", "Stetson_shallow", "Alderdice_deep2")

# Set factor order
site_aggregated$sitelocality <- factor(site_aggregated$sitelocality, levels = siteloc_order_turb)

# Create the site-level plot with updated palette
p_site_w_unknown <- ggplot(site_aggregated, aes(x = sitelocality, y = Relative_Abundance, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack",
           color = "black", size = 0.4, width = 0.85) +
  facet_grid(~ fraction, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = phylum_color_map_w_unknown) +  # Updated palette name
  labs(y = "Relative abundance (%)", x = "Site Locality", 
       title = "Relative Abundance of Top 20 Phyla by Site (Including Unclassified) - 18S") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12, color = "black"),
    axis.text.y = element_text(size = 16, color = "black"),
    axis.title = element_text(size = 17, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 17, face = "bold", margin = margin(t = 0, b = 5)),
    legend.title = element_text(size = 17, face = "bold", color = "black"),
    legend.text = element_text(size = 16, color = "black"),
    legend.position = "right",
    panel.grid = element_blank(),
    panel.spacing = unit(0.5, "cm")
  )

# Save the plot
ggsave(file.path(output_dir, "18S_Top20_Phyla_RelativeAbundance_BySite_WITH_UNCLASSIFIED.pdf"), 
       p_site_w_unknown, width = 14, height = 10)

cat("\n✓ Site-aggregated plot saved to:", 
    file.path(output_dir, "18S_Top20_Phyla_RelativeAbundance_BySite_WITH_UNCLASSIFIED.pdf"), "\n")

# Display the plot
print(p_site_w_unknown)
```





=====================================
----------- ANCOMBC------------------
=====================================
# Load libraries
```{r load ANCOMBC2 libraries}
library(phyloseq)
library(ANCOMBC)
library(tidyverse)
library(conflicted)
library(foreach)
library(rngtools)
library(reshape2)
```


# Create phyloseq object using raw counts
biotic_clean = asv table
metazoo_taxatab rows = ASVs

```{r create phyloseq object using raw counts}
# Make sure ASV table is correct dimensions and a matrix
biotic_clean_18S <- as.matrix(biotic_clean_18S)
str(biotic_clean_18S)

# Create phyloseq object
ancom_physeq <- phyloseq(otu_table(biotic_clean_18S, taxa_are_rows = TRUE),
                                sample_data(metadata_18S),
                                tax_table(metazoo_tax_tab))
# Export phyloseq object
     save(ancom_physeq, file = "/Users/nicolepittoors/Documents/Documents - biscaya/Bioinformatics/ARMS_18S_LULU/FINAL_18S_Pipeline/Taxa_Output_18S/ancombc_raw_counts_physeq_18S.RData")
```

#Set prv_cut = 0.1. 0.1667 may be too conservative. 
Use raw count data
Since size fraction is a group, that would be 1/6 of the samples = 0.1667
Using psuedo_sens = TRUE
Pseudo-counts are small values (0.1, 0.5, 1) added to zero counts before log transformation, 
but different pseudo-count choices can lead to false positives. The sensitivity analysis 
(pseudo_sens = TRUE) re-runs the analysis with multiple pseudo-count values to identify 
which significant results are robust across different choices versus those that are artifacts 
of the specific pseudo-count used. This ensures our differentially abundant taxa findings are 
reliable and not dependent on arbitrary data preprocessing decisions.
```{r fraction check ANCOMBC}
fraction_ancombc <- ancombc2(data = ancom_physeq, 
                            fix_formula = "fraction",
                             global = TRUE, 
                             p_adj_method = "BH", 
                             prv_cut = 0.10, 
                             group = "fraction",
                            pseudo_sens = TRUE)

#Check for results
print(fraction_ancombc$res_global)
```


# 1) Exctract top 20 most significant ASVs
Extract the results from the ANCOMBC2 output, sort them by their p-values (or adjusted p-values), and then select the top 20 ASVs
```{r Extract top 20 most significant ASVs}
# Extract results from ANCOMBC2 output
ancom_res <- fraction_ancombc$res_global

# Sort the ASVs by the p-value or adjusted p-value (q_val) and select the top 20
top20_asvs <- ancom_res %>%
  dplyr::arrange(q_val) %>%
  head(20) %>%
  pull(taxon)  # Extract the ASV IDs

# Print the top 20 ASVs
print(top20_asvs)
```


2) Map the ASVs to their scientific name
```{r map ASVs to scientific names}
# Extract the taxonomic table from the phyloseq object
tax_table <- as.data.frame(tax_table(ancom_physeq))

# Filter the taxonomic table to only include the top 20 ASVs
top20_taxonomy <- tax_table[rownames(tax_table) %in% top20_asvs, ]

# Melt the phyloseq object (this already includes taxonomy)
physeq_melted <- psmelt(ancom_physeq)

# Filter to only include the top 20 ASVs 
physeq_melted_top20 <- physeq_melted[physeq_melted$OTU %in% top20_asvs, ]

# Check the result
colnames(physeq_melted_top20)  # Should have clean column names

# Check cleaned column names
colnames(physeq_melted_clean)

# Remove ALL duplicate columns, keeping only the first occurrence
physeq_melted_clean <- physeq_melted_clean[, !duplicated(colnames(physeq_melted_clean))]

# Now run your analysis
physeq_melted_clean <- physeq_melted_clean %>%
  dplyr::filter(OTU %in% top20_asvs) %>%
  dplyr::arrange(fraction, Sample) %>%
  dplyr::mutate(Sample = factor(Sample, levels = unique(Sample)))

colnames(physeq_melted_clean)
```


3) Plot the abundance of the top 20 ASVs
```{r Plot the abundance of the top 20 ASVs}
# Check if 'fraction' column exists
if (!"fraction" %in% colnames(physeq_melted_clean)) {
  stop("The 'fraction' column is not found in the sample data.")
}

# Order the samples by 'fraction' and 'Sample'
physeq_melted_frac <- physeq_melted_clean %>%
  dplyr::filter(OTU %in% top20_asvs) %>%  # Filter only top 20 ASVs
  dplyr::mutate(Sample = factor(Sample, levels = unique(Sample[order(fraction, Sample)])))

# Create the plot with custom colors
fraction_plot1 <- ggplot(physeq_melted_frac, aes(x = Sample, y = Abundance, fill = family)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  labs(title = "Top 20 Significant ASVs Grouped by Fraction-18S", 
       x = "Sample", 
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_grid(~ fraction, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c20)  

# Print the plot
print(fraction_plot1)

# Save
ggsave("/Users/nicolepittoors/Documents/Documents - biscaya/Bioinformatics/ARMS_18S_LULU/FINAL_18S_Pipeline/Taxa_Output_18S/ANCOMBC_abun_fraction_18S.pdf",
      plot = last_plot(), 
       device = "pdf", 
       scale = 1, 
       width = 11, 
       height = 8.5, 
       units = "in", 
       dpi = 300)
```


# Export .CSV of mapped unique families
```{r export csv of unique families}
tax_subset_18S <- tax_table[, c("phylum", "class", "order", "family")]
unique_families_18S <- unique(tax_subset_18S)

# Remove any rows where Family is NA
unique_families_18S <- unique_families_18S[order(unique_families_18S$phylum, 
                                          unique_families_18S$class,
                                          unique_families_18S$order,
                                          unique_families_18S$family), ]


unique_families_clean_18S <- unique_families_18S[!grepl("[^[:alnum:]]$", unique_families_18S$family), ]

# Remove taxonomic errors
taxonomic_errors <- c("Annelida", "Nemertea", "Phoronida", "Phyllodocida", 
                      "Chromadorea", "Pteropoda")

unique_families_metazoa_only <- unique_families_clean_18S[
  !unique_families_clean_18S$family %in% taxonomic_errors, ]

write.csv(unique_families_metazoa_only, 
          file = "/Users/nicolepittoors/Documents/Documents - biscaya/Bioinformatics/ARMS_18S_LULU/FINAL_18S_Pipeline/Taxa_Output_18S/unique_families_18S_clean.csv", 
          row.names = FALSE)
```


# Subset sessile and motile fractions
```{r subset fractions}
# Subset phyloseq objects - combine 100 and 500 as motile
sessile_physeq <- subset_samples(ancom_physeq, fraction == "sessile")
motile_physeq <- subset_samples(ancom_physeq, fraction %in% c("100", "500")) 

# Check sample sizes
cat("Sessile samples:", nsamples(sessile_physeq), "\n")
cat("Motile samples (100 + 500):", nsamples(motile_physeq), "\n")

# Check the breakdown of motile samples
motile_breakdown <- table(sample_data(motile_physeq)$fraction)
cat("Motile fraction breakdown:\n")
print(motile_breakdown)

# Remove taxa with zero abundance after subsetting
sessile_physeq <- prune_taxa(taxa_sums(sessile_physeq) > 0, sessile_physeq)
motile_physeq <- prune_taxa(taxa_sums(motile_physeq) > 0, motile_physeq)

cat("\n=== SESSILE FRACTION ANALYSIS ===\n")
sessile_ancombc <- ancombc2(
  data = sessile_physeq,
  tax_level = "phylum",
  fix_formula = "turbidity_std_rank + depth",
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.10
)

cat("\n=== MOTILE FRACTIONS (100 + 500) ANALYSIS ===\n")
motile_ancombc <- ancombc2(
  data = motile_physeq,
  tax_level = "phylum",
  fix_formula = "turbidity_std_rank + depth",
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.10
)

# Extract sessile results
sessile_turb <- sessile_ancombc$res %>%
  dplyr::filter(q_turbidity_std_rank < 0.05) %>%
  dplyr::arrange(q_turbidity_std_rank) %>%
  dplyr::select(taxon, lfc_turbidity_std_rank, p_turbidity_std_rank, q_turbidity_std_rank)

sessile_depth <- sessile_ancombc$res %>%
  dplyr::filter(q_depth < 0.05) %>%
  dplyr::arrange(q_depth) %>%
  dplyr::select(taxon, lfc_depth, p_depth, q_depth)

cat("Sessile - Significant phyla for turbidity:\n")
print(sessile_turb)
cat("Sessile - Significant phyla for depth:\n")
print(sessile_depth)

# Extract motile results
motile_turb <- motile_ancombc$res %>%
  dplyr::filter(q_turbidity_std_rank < 0.05) %>%
  dplyr::arrange(q_turbidity_std_rank) %>%
  dplyr::select(taxon, lfc_turbidity_std_rank, p_turbidity_std_rank, q_turbidity_std_rank)

motile_depth <- motile_ancombc$res %>%
  dplyr::filter(q_depth < 0.05) %>%
  dplyr::arrange(q_depth) %>%
  dplyr::select(taxon, lfc_depth, p_depth, q_depth)

cat("Motile - Significant phyla for turbidity:\n")
print(motile_turb)
cat("Motile - Significant phyla for depth:\n")
print(motile_depth)
```


# Test for significant differences in Phyla abundances across turbidty and depth gradients in ALL SAMPLES
```{r phlya differences depth/turb all samples}
## CHECK SPECIFICALLY FOR RHODOPHYTA
cat("\n=== RHODOPHYTA SPECIFIC RESULTS ===\n")

# Sessile Rhodophyta
sessile_rhodo <- sessile_ancombc$res %>%
  dplyr::filter(taxon == "Rhodophyta") %>%
  dplyr::select(taxon, lfc_turbidity_std_rank, q_turbidity_std_rank, lfc_depth, q_depth)

# Motile Rhodophyta  
motile_rhodo <- motile_ancombc$res %>%
  dplyr::filter(taxon == "Rhodophyta") %>%
  dplyr::select(taxon, lfc_turbidity_std_rank, q_turbidity_std_rank, lfc_depth, q_depth)

cat("Sessile Rhodophyta results:\n")
print(sessile_rhodo)
cat("Motile Rhodophyta results:\n")
print(motile_rhodo)
```



# ----------------------- ANCOM-BC2 FRACTION, SITE, DEPTH + TURB ------------


#  ------ 1. ANCOMBC phyla and families associated with each FRACTION ------
Which taxa are enriched in each fraction
Size-based community structure
```{r fraction effects, phylum fam}
# ========== FRACTION EFFECTS ANALYSIS  ========

# Phylum level - fraction effects
fraction_P_ancombc <- ancombc2(
  data = ancom_physeq,
  tax_level = "phylum",
  fix_formula = "fraction",  # Testing 100µm vs 500µm vs sessile
  global = TRUE,             # Global test across fractions
  pairwise = TRUE,           # Which specific fractions differ
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "fraction",
  pseudo_sens = TRUE
)

# Family level - fraction effects  
fraction_F_ancombc <- ancombc2(
  data = ancom_physeq,
  tax_level = "family",
  fix_formula = "fraction",
  global = TRUE,
  pairwise = TRUE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "fraction",
  pseudo_sens = TRUE
)

# Function to extract and format fraction results
extract_fraction_results <- function(ancombc_result, tax_level) {
  
  # Global results (which taxa differ between fractions)
  global_res <- ancombc_result$res_global
  global_res$Analysis <- paste0("Fraction_Effects_", tax_level)
  global_res$Test_Type <- "Global"
  
  # Pairwise results (which specific fraction pairs differ)
  pairwise_res <- ancombc_result$res_pair
  pairwise_res$Analysis <- paste0("Fraction_Effects_", tax_level)
  pairwise_res$Test_Type <- "Pairwise"
  
  return(list(global = global_res, pairwise = pairwise_res))
}

# Extract fraction results
fraction_phylum_results <- extract_fraction_results(fraction_P_ancombc, "Phylum")
fraction_family_results <- extract_fraction_results(fraction_F_ancombc, "Family")

# Export fraction results
write.csv(fraction_phylum_results$global, file.path(output_dir, "fraction_phylum_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(fraction_phylum_results$pairwise, file.path(output_dir, "fraction_phylum_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(fraction_family_results$global, file.path(output_dir, "fraction_family_global_results_18S.csv"), row.names = FALSE)
write.csv(fraction_family_results$pairwise, file.path(output_dir, "fraction_family_pairwise_results_18S.csv"), row.names = FALSE)
```




# ------ 2. ANCOMBC phyla and families associated with each SITE -------

Answers the questions:
Which taxa vary significantly across the 12 sites
Unmeasured environmental factors creating site-specific communities
Geographic patterns in community structure
```{r site, phylum and fam}
# 1. All samples - Phylum level
site_global_P_ancombc <- ancombc2(
  data = ancom_physeq,
  tax_level = "phylum",
  fix_formula = "sitelocality",  # Categorical site variable
  global = TRUE,               # Global test across all sites
  pairwise = TRUE,             # Which specific sites differ
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "sitelocality",      # Grouping variable for global test
  pseudo_sens = TRUE
)

# 2. All samples - Family level  
site_global_F_ancombc <- ancombc2(
  data = ancom_physeq,
  tax_level = "family",
  fix_formula = "sitelocality",
  global = TRUE,
  pairwise = TRUE,
  p_adj_method = "BH", 
  prv_cut = 0.1,
  group = "sitelocality",
  pseudo_sens = TRUE
)

# 3. Motile site effects - Phylum
motile_site_ancombc <- ancombc2(
  data = motile_physeq,
  tax_level = "phylum",
  fix_formula = "sitelocality",
  global = TRUE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "sitelocality",
  pseudo_sens = TRUE
)

# 4. Motile site effects - Family
motile_site_FAMancombc <- ancombc2(
  data = motile_physeq,
  tax_level = "family",
  fix_formula = "sitelocality",
  global = TRUE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "sitelocality",
  pseudo_sens = TRUE
)

# 5. Sessile site effects - Phylum
sessile_site_ancombc <- ancombc2(
  data = sessile_physeq,
  tax_level = "phylum",
  fix_formula = "sitelocality",
  global = TRUE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "sitelocality",
  pseudo_sens = TRUE
)

# 6. Sessile site effects - Family
sessile_site_FAMancombc <- ancombc2(
  data = sessile_physeq,
  tax_level = "family",
  fix_formula = "sitelocality",
  global = TRUE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  group = "sitelocality",
  pseudo_sens = TRUE
)

# EXPORT SITE EFFECTS RESULTS

# Function to extract and format site results
extract_site_results <- function(ancombc_result, analysis_name) {
  
  # Global results
  global_res <- ancombc_result$res_global
  global_res$Analysis <- analysis_name
  global_res$Test_Type <- "Global"
  
  # Pairwise results (if available)
  if("res_pair" %in% names(ancombc_result)) {
    pairwise_res <- ancombc_result$res_pair  
    pairwise_res$Analysis <- analysis_name
    pairwise_res$Test_Type <- "Pairwise"
  } else {
    pairwise_res <- data.frame()
  }
  
  return(list(global = global_res, pairwise = pairwise_res))
}

# Extract all site results
site_all_phylum <- extract_site_results(site_global_P_ancombc, "All_Samples_Phylum")
site_all_family <- extract_site_results(site_global_F_ancombc, "All_Samples_Family")
site_motile_phylum <- extract_site_results(motile_site_ancombc, "Motile_Phylum")
site_motile_family <- extract_site_results(motile_site_FAMancombc, "Motile_Family")
site_sessile_phylum <- extract_site_results(sessile_site_ancombc, "Sessile_Phylum")
site_sessile_family <- extract_site_results(sessile_site_FAMancombc, "Sessile_Family")

# Export individual site global results
write.csv(site_all_phylum$global, file.path(output_dir, "site_all_phylum_global_results_18S.csv"), row.names = FALSE)
write.csv(site_all_family$global, file.path(output_dir, "site_all_family_global_results_18S.csv"), row.names = FALSE)
write.csv(site_motile_phylum$global, file.path(output_dir, "site_motile_phylum_global_results_18S.csv"), row.names = FALSE)
write.csv(site_motile_family$global, file.path(output_dir, "site_motile_family_global_results_18S.csv"), row.names = FALSE)
write.csv(site_sessile_phylum$global, file.path(output_dir, "site_sessile_phylum_global_results_18S.csv"), row.names = FALSE)
write.csv(site_sessile_family$global, file.path(output_dir, "site_sessile_family_global_results_18S.csv"), row.names = FALSE)

# Export individual site pairwise results (if available)
write.csv(site_all_phylum$pairwise, file.path(output_dir, "site_all_phylum_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(site_all_family$pairwise, file.path(output_dir, "site_all_family_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(site_motile_phylum$pairwise, file.path(output_dir, "site_motile_phylum_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(site_motile_family$pairwise, file.path(output_dir, "site_motile_family_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(site_sessile_phylum$pairwise, file.path(output_dir, "site_sessile_phylum_pairwise_results_18S.csv"), row.names = FALSE)
write.csv(site_sessile_family$pairwise, file.path(output_dir, "site_sessile_family_pairwise_results_18S.csv"), row.names = FALSE)

# Combine all global site results for comparison
all_site_global <- rbind(
  site_all_phylum$global,
  site_all_family$global,
  site_motile_phylum$global,
  site_motile_family$global,
  site_sessile_phylum$global,
  site_sessile_family$global
)

write.csv(all_site_global, file.path(output_dir, "all_site_global_results_combined_18S.csv"), row.names = FALSE)

# Create summary of significant results
create_summary <- function(global_results, analysis_name) {
  n_significant <- sum(global_results$diff_abn, na.rm = TRUE)
  n_total <- nrow(global_results)
  return(data.frame(
    Analysis = analysis_name,
    Total_Taxa = n_total,
    Significant_Taxa = n_significant,
    Percent_Significant = round((n_significant/n_total)*100, 1)
  ))
}

# Create summary table
summary_table <- rbind(
  create_summary(site_all_phylum$global, "All_Samples_Phylum"),
  create_summary(site_all_family$global, "All_Samples_Family"),
  create_summary(site_motile_phylum$global, "Motile_Phylum"),
  create_summary(site_motile_family$global, "Motile_Family"),
  create_summary(site_sessile_phylum$global, "Sessile_Phylum"),
  create_summary(site_sessile_family$global, "Sessile_Family"),
  create_summary(fraction_phylum_results$global, "Fraction_Effects_Phylum"),
  create_summary(fraction_family_results$global, "Fraction_Effects_Family")
)

write.csv(summary_table, file.path(output_dir, "all_analyses_summary_table_18S.csv"), row.names = FALSE)

# Print summary
print(summary_table)
```


```{r}
# Set preference for dplyr functions
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)

# Get all LFC columns (both individual sites and pairwise comparisons)
lfc_cols <- grep("^lfc_sitelocality", colnames(site_all_phylum$pairwise), value = TRUE)
p_cols <- grep("^p_sitelocality", colnames(site_all_phylum$pairwise), value = TRUE)

# Summary of effect sizes across all comparisons
all_lfc_values <- unlist(site_all_phylum$pairwise[, lfc_cols])
cat("Summary of all log-fold changes across site comparisons:\n")
print(summary(abs(all_lfc_values)))

# Find the largest effects for each taxon
max_effects_per_taxon <- site_all_phylum$pairwise %>%
  rowwise() %>%
  dplyr::mutate(
    max_lfc = max(abs(c_across(all_of(lfc_cols))), na.rm = TRUE),
    min_pval = min(c_across(all_of(p_cols)), na.rm = TRUE)
  ) %>%
  dplyr::select(taxon, max_lfc, min_pval) %>%
  arrange(desc(max_lfc))

cat("Taxa with largest site effects:\n")
print(max_effects_per_taxon)

# Count significant raw p-values (before correction)
raw_sig_count <- site_all_phylum$pairwise %>%
  dplyr::select(taxon, all_of(p_cols)) %>%
  rowwise() %>%
  dplyr::mutate(
    significant_raw_comparisons = sum(c_across(all_of(p_cols)) < 0.05, na.rm = TRUE),
    total_comparisons = length(p_cols)
  ) %>%
  dplyr::select(taxon, significant_raw_comparisons, total_comparisons) %>%
  arrange(desc(significant_raw_comparisons))

cat("Raw significant comparisons per taxon (before multiple testing correction):\n")
print(raw_sig_count)
```


# ------ 3. ANCOMBC phyla and families associated with DEPTH AND TURB ------
```{r depth and turb, phylum fam}
# ============ ENVIRONMENTAL EFFECTS ANALYSIS: DEPTH + TURBIDITY ===============
# Testing across All Samples, Motile Fractions, and Sessile Fractions

# Create ARMS ID for random effects (if needed later)
sample_data(ancom_physeq)$ARMS_corrected <- paste(
  sample_data(ancom_physeq)$sitelocality,
  sample_data(ancom_physeq)$ARMS_ID_corrected,
  sep = "_"
)

# ============== 1. ANCOM-BC2 ANALYSES: Environmental Effects (Depth + Turbidity) ==============

cat("Running ANCOM-BC2 environmental analyses...\n")

# 1/6 All Samples - Phylum Level
env_phylum_all <- ancombc2(
  data = ancom_physeq,
  tax_level = "phylum",
  fix_formula = "depth + turbidity_std_rank",
  rand_formula = NULL,
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  pseudo_sens = TRUE
)

# 2/6 All Samples - Family Level  
env_family_all <- ancombc2(
  data = ancom_physeq,
  tax_level = "family",
  fix_formula = "depth + turbidity_std_rank",
  rand_formula = NULL,
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  pseudo_sens = TRUE
)

# 3/6 Motile Fractions - Phylum Level
env_phylum_motile <- ancombc2(
  data = motile_physeq,
  tax_level = "phylum",
  fix_formula = "depth + turbidity_std_rank",
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  pseudo_sens = TRUE
)

# 4/6 Motile Fractions - Family Level
env_family_motile <- ancombc2(
  data = motile_physeq,
  tax_level = "family",
  fix_formula = "depth + turbidity_std_rank",
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  pseudo_sens = TRUE
)

# 5/6 Sessile Fraction - Phylum Level
env_phylum_sessile <- ancombc2(
  data = sessile_physeq,
  tax_level = "phylum",
  fix_formula = "depth + turbidity_std_rank",
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  pseudo_sens = TRUE
)

# 6/6 Sessile Fraction - Family Level
env_family_sessile <- ancombc2(
  data = sessile_physeq,
  tax_level = "family",
  fix_formula = "depth + turbidity_std_rank",
  global = FALSE,
  p_adj_method = "BH",
  prv_cut = 0.1,
  pseudo_sens = TRUE
)

cat("All ANCOM-BC2 analyses completed.\n")


# ============== 2. EXTRACT RESULTS =============

# Extract results objects
env_results_phylum_all <- env_phylum_all$res
env_results_family_all <- env_family_all$res
env_results_phylum_motile <- env_phylum_motile$res
env_results_family_motile <- env_family_motile$res
env_results_phylum_sessile <- env_phylum_sessile$res
env_results_family_sessile <- env_family_sessile$res

# ============= 3. FUNCTION TO EXTRACT SIGNIFICANT TAXA =========
extract_significant_environmental_taxa <- function(results, analysis_name, fraction_type, tax_level) {
  
  # Extract significant taxa for depth
  depth_sig_indices <- which(results$diff_depth == TRUE)
  turbidity_sig_indices <- which(results$diff_turbidity_std_rank == TRUE)
  
  # Create depth results dataframe
  if(length(depth_sig_indices) > 0) {
    depth_results <- data.frame(
      Analysis = analysis_name,
      Fraction = fraction_type,
      Taxonomic_Level = tax_level,
      Environmental_Variable = "Depth",
      Taxon = results$taxon[depth_sig_indices],
      Log_Fold_Change = round(results$lfc_depth[depth_sig_indices], 4),
      P_value = results$p_depth[depth_sig_indices],
      Q_value = results$q_depth[depth_sig_indices],
      Robust = results$passed_ss_depth[depth_sig_indices],
      Direction = ifelse(results$lfc_depth[depth_sig_indices] > 0, "Increases_with_depth", "Decreases_with_depth"),
      stringsAsFactors = FALSE
    )
  } else {
    depth_results <- data.frame()
  }
  
  # Create turbidity results dataframe
  if(length(turbidity_sig_indices) > 0) {
    turbidity_results <- data.frame(
      Analysis = analysis_name,
      Fraction = fraction_type,
      Taxonomic_Level = tax_level,
      Environmental_Variable = "Turbidity",
      Taxon = results$taxon[turbidity_sig_indices],
      Log_Fold_Change = round(results$lfc_turbidity_std_rank[turbidity_sig_indices], 4),
      P_value = results$p_turbidity_std_rank[turbidity_sig_indices],
      Q_value = results$q_turbidity_std_rank[turbidity_sig_indices],
      Robust = results$passed_ss_turbidity_std_rank[turbidity_sig_indices],
      Direction = ifelse(results$lfc_turbidity_std_rank[turbidity_sig_indices] > 0, "Increases_with_turbidity", "Decreases_with_turbidity"),
      stringsAsFactors = FALSE
    )
  } else {
    turbidity_results <- data.frame()
  }
  
  # Combine and return
  combined_results <- rbind(depth_results, turbidity_results)
  return(combined_results)
}

# ==============  4. EXTRACT SIGNIFICANT RESULTS FOR ALL ANALYSES ================

cat("Extracting significant results...\n")

# Extract significant results
sig_env_phylum_all <- extract_significant_environmental_taxa(env_results_phylum_all, "Environmental_Effects", "All_Samples", "Phylum")
sig_env_family_all <- extract_significant_environmental_taxa(env_results_family_all, "Environmental_Effects", "All_Samples", "Family")
sig_env_phylum_motile <- extract_significant_environmental_taxa(env_results_phylum_motile, "Environmental_Effects", "Motile", "Phylum")
sig_env_family_motile <- extract_significant_environmental_taxa(env_results_family_motile, "Environmental_Effects", "Motile", "Family")
sig_env_phylum_sessile <- extract_significant_environmental_taxa(env_results_phylum_sessile, "Environmental_Effects", "Sessile", "Phylum")
sig_env_family_sessile <- extract_significant_environmental_taxa(env_results_family_sessile, "Environmental_Effects", "Sessile", "Family")

# Combine all significant results
all_environmental_significant <- rbind(
  sig_env_phylum_all,
  sig_env_family_all,
  sig_env_phylum_motile,
  sig_env_family_motile,
  sig_env_phylum_sessile,
  sig_env_family_sessile
)

# ===================  5. EXPORT RESULTS TO CSV ===================

cat("Exporting results to CSV files...\n")

# Export full results (all taxa, not just significant)
write.csv(env_results_phylum_all, file.path(output_dir, "environmental_phylum_all_samples_full_18S.csv"), row.names = FALSE)
write.csv(env_results_family_all, file.path(output_dir, "environmental_family_all_samples_full_18S.csv"), row.names = FALSE)
write.csv(env_results_phylum_motile, file.path(output_dir, "environmental_phylum_motile_full_18S.csv"), row.names = FALSE)
write.csv(env_results_family_motile, file.path(output_dir, "environmental_family_motile_full_18S.csv"), row.names = FALSE)
write.csv(env_results_phylum_sessile, file.path(output_dir, "environmental_phylum_sessile_full_18S.csv"), row.names = FALSE)
write.csv(env_results_family_sessile, file.path(output_dir, "environmental_family_sessile_full_18S.csv"), row.names = FALSE)

# Export significant results only
write.csv(sig_env_phylum_all, file.path(output_dir, "environmental_phylum_all_samples_significant_18S.csv"), row.names = FALSE)
write.csv(sig_env_family_all, file.path(output_dir, "environmental_family_all_samples_significant_18S.csv"), row.names = FALSE)
write.csv(sig_env_phylum_motile, file.path(output_dir, "environmental_phylum_motile_significant_18S.csv"), row.names = FALSE)
write.csv(sig_env_family_motile, file.path(output_dir, "environmental_family_motile_significant_18S.csv"), row.names = FALSE)
write.csv(sig_env_phylum_sessile, file.path(output_dir,  "environmental_phylum_sessile_significant_18S.csv"), row.names = FALSE)
write.csv(sig_env_family_sessile, file.path(output_dir, "environmental_family_sessile_significant_18S.csv"), row.names = FALSE)

# Export combined significant results
write.csv(all_environmental_significant, file.path(output_dir, "all_environmental_significant_results_18S.csv"), row.names = FALSE)


# =================== 6. CREATE SUMMARY TABLE ===================

env_summary_table <- data.frame(
  Analysis = c("All_Samples_Phylum", "All_Samples_Family", "Motile_Phylum", "Motile_Family", "Sessile_Phylum", "Sessile_Family"),
  Total_Taxa_Tested = c(
    nrow(env_results_phylum_all), nrow(env_results_family_all),
    nrow(env_results_phylum_motile), nrow(env_results_family_motile),
    nrow(env_results_phylum_sessile), nrow(env_results_family_sessile)
  ),
  Significant_Depth = c(
    sum(env_results_phylum_all$diff_depth, na.rm = TRUE),
    sum(env_results_family_all$diff_depth, na.rm = TRUE),
    sum(env_results_phylum_motile$diff_depth, na.rm = TRUE),
    sum(env_results_family_motile$diff_depth, na.rm = TRUE),
    sum(env_results_phylum_sessile$diff_depth, na.rm = TRUE),
    sum(env_results_family_sessile$diff_depth, na.rm = TRUE)
  ),
  Significant_Turbidity = c(
    sum(env_results_phylum_all$diff_turbidity_std_rank, na.rm = TRUE),
    sum(env_results_family_all$diff_turbidity_std_rank, na.rm = TRUE),
    sum(env_results_phylum_motile$diff_turbidity_std_rank, na.rm = TRUE),
    sum(env_results_family_motile$diff_turbidity_std_rank, na.rm = TRUE),
    sum(env_results_phylum_sessile$diff_turbidity_std_rank, na.rm = TRUE),
    sum(env_results_family_sessile$diff_turbidity_std_rank, na.rm = TRUE)
  ),
  Total_Significant = c(
    nrow(sig_env_phylum_all), nrow(sig_env_family_all),
    nrow(sig_env_phylum_motile), nrow(sig_env_family_motile),
    nrow(sig_env_phylum_sessile), nrow(sig_env_family_sessile)
  )
)

write.csv(env_summary_table, file.path(output_dir, "environmental_analysis_summary.csv"), row.names = FALSE)


# =================== 7. PRINT RESULTS SUMMARY ===================

cat("\n=== ENVIRONMENTAL ANALYSIS SUMMARY ===\n")
print(env_summary_table)

cat("\n=== SIGNIFICANT RESULTS PREVIEW ===\n")
if(nrow(all_environmental_significant) > 0) {
  print(head(all_environmental_significant, 15))
  cat("\nTotal significant environmental effects:", nrow(all_environmental_significant), "\n")
  cat("Robust effects (passed sensitivity analysis):", sum(all_environmental_significant$Robust), "\n")
} else {
  cat("No significant results found\n")
}
```

# ------ 4. Tables with statistics ------
```{r export statistics}
# Function to extract detailed environmental effects
extract_detailed_environmental_effects <- function(results, analysis_name, fraction_name, tax_level) {
  
  # Extract depth effects (significant only)
  depth_significant <- which(results$diff_depth == TRUE)
  
  if(length(depth_significant) > 0) {
    depth_data <- data.frame(
      Analysis = analysis_name,
      Fraction = fraction_name,
      Taxonomic_Level = tax_level,
      Environmental_Variable = "Depth",
      Taxon = results$taxon[depth_significant],
      Log_Fold_Change = round(results$lfc_depth[depth_significant], 4),
      P_value = format(results$p_depth[depth_significant], scientific = TRUE, digits = 3),
      Q_value = format(results$q_depth[depth_significant], scientific = TRUE, digits = 3),
      Robust = results$passed_ss_depth[depth_significant],
      Direction = ifelse(results$lfc_depth[depth_significant] > 0, "Increases", "Decreases"),
      stringsAsFactors = FALSE
    )
  } else {
    depth_data <- data.frame()
  }
  
  # Extract turbidity effects (significant only)
  turbidity_significant <- which(results$diff_turbidity_std_rank == TRUE)
  
  if(length(turbidity_significant) > 0) {
    turbidity_data <- data.frame(
      Analysis = analysis_name,
      Fraction = fraction_name,
      Taxonomic_Level = tax_level,
      Environmental_Variable = "Turbidity",
      Taxon = results$taxon[turbidity_significant],
      Log_Fold_Change = round(results$lfc_turbidity_std_rank[turbidity_significant], 4),
      P_value = format(results$p_turbidity_std_rank[turbidity_significant], scientific = TRUE, digits = 3),
      Q_value = format(results$q_turbidity_std_rank[turbidity_significant], scientific = TRUE, digits = 3),
      Robust = results$passed_ss_turbidity_std_rank[turbidity_significant],
      Direction = ifelse(results$lfc_turbidity_std_rank[turbidity_significant] > 0, "Increases", "Decreases"),
      stringsAsFactors = FALSE
    )
  } else {
    turbidity_data <- data.frame()
  }
  
  # Combine depth and turbidity results
  combined_data <- rbind(depth_data, turbidity_data)
  
  return(combined_data)
}

# Extract detailed results from all environmental analyses
detailed_results_list <- list(
  extract_detailed_environmental_effects(env_results_phylum, "Environmental_Effects", "All_Samples", "Phylum"),
  extract_detailed_environmental_effects(env_results, "Environmental_Effects", "All_Samples", "Family"),
  extract_detailed_environmental_effects(motile_PHY_env_results, "Environmental_Effects", "Motile", "Phylum"),
  extract_detailed_environmental_effects(motile_FAM_env_results, "Environmental_Effects", "Motile", "Family"),
  extract_detailed_environmental_effects(sessile_PHY_env_results, "Environmental_Effects", "Sessile", "Phylum"),
  extract_detailed_environmental_effects(sessile_FAM_env_results, "Environmental_Effects", "Sessile", "Family")
)

# Combine all results into one comprehensive table
detailed_environmental_table <- do.call(rbind, detailed_results_list)

# Sort by fraction, taxonomic level, environmental variable, and q-value
detailed_environmental_table <- detailed_environmental_table[
  order(detailed_environmental_table$Fraction, 
        detailed_environmental_table$Taxonomic_Level,
        detailed_environmental_table$Environmental_Variable,
        as.numeric(detailed_environmental_table$Q_value)), 
]

# Add row numbers for reference
detailed_environmental_table$Entry_ID <- 1:nrow(detailed_environmental_table)

# Reorder columns for clarity
detailed_environmental_table <- detailed_environmental_table[, c("Entry_ID", "Analysis", "Fraction", "Taxonomic_Level", 
                                                               "Environmental_Variable", "Taxon", "Log_Fold_Change", 
                                                               "P_value", "Q_value", "Robust", "Direction")]

# Export the detailed table
write.csv(detailed_environmental_table, "Detailed_Environmental_Effects_Table.csv", row.names = FALSE)

# Create a version with only robust results
robust_only_table <- detailed_environmental_table[detailed_environmental_table$Robust == TRUE, ]

write.csv(robust_only_table, "Robust_Environmental_Effects_Table.csv", row.names = FALSE)

# Print preview of the table
print("=== DETAILED ENVIRONMENTAL EFFECTS TABLE (First 20 rows) ===")
print(head(detailed_environmental_table, 20))

# Print summary statistics
cat("\n=== SUMMARY STATISTICS ===\n")
cat("Total significant environmental effects:", nrow(detailed_environmental_table), "\n")
cat("Robust effects (passed sensitivity analysis):", sum(detailed_environmental_table$Robust), "\n")
cat("Depth effects:", sum(detailed_environmental_table$Environmental_Variable == "Depth"), "\n")
cat("Turbidity effects:", sum(detailed_environmental_table$Environmental_Variable == "Turbidity"), "\n")

# Create summary by fraction and taxonomic level
summary_by_analysis <- detailed_environmental_table %>%
  group_by(Fraction, Taxonomic_Level, Environmental_Variable) %>%
  dplyr::summarise(
    Total_Effects = n(),
    Robust_Effects = sum(Robust),
    Increases = sum(Direction == "Increases"),
    Decreases = sum(Direction == "Decreases"),
    .groups = "drop"
  )

print("\n=== SUMMARY BY ANALYSIS TYPE ===")
print(summary_by_analysis)

write.csv(summary_by_analysis, "Environmental_Effects_Summary_by_Analysis.csv", row.names = FALSE)

# Create a version showing the strongest effects (lowest q-values)
top_effects <- detailed_environmental_table[1:min(20, nrow(detailed_environmental_table)), ]

print("\n=== TOP 20 STRONGEST ENVIRONMENTAL EFFECTS ===")
print(top_effects)

write.csv(top_effects, "Top_Environmental_Effects.csv", row.names = FALSE)
```


# ------ 5. Create visualization for each fraction- EDITING FOR FINAL ------


# Barplots of Phyla abundance across sites, colored by turbidity values (cdivis)
```{r barplots }
# =============================================================================
# ENVIRONMENTAL EFFECTS PLOTS: Phyla by Depth and Turbidity
# Updated with Cividis (Turbidity) and Mako Blue Gradient (Depth)
# =============================================================================

# Load required libraries
#library(ggplot2)
#library(dplyr)
#library(phyloseq)
#library(viridis)
#library(gridExtra)

# Resolve conflicts
#conflicts_prefer(dplyr::mutate)
#conflicts_prefer(dplyr::summarise)
#conflicts_prefer(dplyr::filter)

# Function to create environmental effects barplots with improved color schemes
create_environmental_phylum_barplots <- function(physeq_obj, significant_results, 
                                               env_variable = "Turbidity",
                                               fraction_name = "Motile",
                                               y_margin = 0.2) {
  
  # Filter for the specific environmental variable
  env_data <- significant_results %>% 
    filter(Environmental_Variable == env_variable & Taxonomic_Level == "Phylum")
  
  if(nrow(env_data) == 0) {
    message(paste("No significant", env_variable, "effects found for", fraction_name, "fraction"))
    return(NULL)
  }
  
  # Get phyla of interest
  phyla_of_interest <- env_data$Taxon
  
  # Transform phyloseq data for plotting
  physeq_rel <- transform_sample_counts(physeq_obj, function(x) x/sum(x))
  
  # Melt to long format
  plot_data <- psmelt(physeq_rel) %>%
    filter(phylum %in% phyla_of_interest)
  
  # Calculate mean abundance and error bars by site (aggregate environmental variables)
  summary_data <- plot_data %>%
   group_by(phylum, sitelocality) %>%  # Only group by phylum and site
   dplyr::summarise(
      mean_abundance = mean(Abundance, na.rm = TRUE),
      se = sd(Abundance, na.rm = TRUE) / sqrt(n()),
      # Take mean environmental values for each site
      turbidity_std_rank = mean(turbidity_std_rank, na.rm = TRUE),
      depth = mean(depth, na.rm = TRUE),
      .groups = "drop"
   ) %>%
   arrange(sitelocality)
  
  # Create individual plots for each phylum
  phylum_plots <- list()
  
  for(p in phyla_of_interest) {
    # Get data for this phylum
    p_data <- summary_data %>% filter(phylum == p)
    
    # Get effect information
    p_effect <- env_data %>% filter(Taxon == p)
    
    # Create effect direction and significance label
    direction <- ifelse(p_effect$Log_Fold_Change > 0, "↑", "↓")
    robust_status <- ifelse(p_effect$Robust, "**", "*")
    effect_label <- paste0(direction, " ", env_variable, " ", robust_status)
    
    # Determine plot limits
    plot_max_y <- max(p_data$mean_abundance + p_data$se, na.rm = TRUE) * (1 + y_margin)
    
    # Create the barplot
    if(env_variable == "Turbidity") {
      p1 <- ggplot(p_data, aes(x = reorder(sitelocality, turbidity_std_rank), 
                               y = mean_abundance, 
                               fill = turbidity_std_rank)) +
        geom_col(color = "black", size = 0.3) +
        geom_errorbar(aes(ymin = mean_abundance - se, ymax = mean_abundance + se),
                      width = 0.4, size = 0.5) +
        scale_fill_viridis_c(
          option = "cividis",  # Cividis color scale (blue to yellow)
          name = "Turbidity\n(Std Rank)",
          direction = 1  # Higher values = more yellow
        ) +
        labs(
          title = paste0(p, " (", fraction_name, " Fraction)"),
          subtitle = paste0("Effect: ", effect_label, 
                          " (LFC = ", round(p_effect$Log_Fold_Change, 3), 
                          ", q = ", format(p_effect$Q_value, scientific = TRUE, digits = 2), ")"),
          x = "Site (ordered by turbidity)",
          y = "Relative Abundance"
        )
    } else {
      # For depth effects, use mako gradient (darker blue = deeper)
      p1 <- ggplot(p_data, aes(x = reorder(sitelocality, depth), 
                               y = mean_abundance, 
                               fill = depth)) +
        geom_col(color = "black", size = 0.3) +
        geom_errorbar(aes(ymin = mean_abundance - se, ymax = mean_abundance + se),
                      width = 0.4, size = 0.5) +
        scale_fill_viridis_c(
          option = "mako",  # Mako color scale
          name = "Depth\n(meters)",
          direction = -1  # Darker blue = deeper (higher depth values)
        ) +
        labs(
          title = paste0(p, " (", fraction_name, " Fraction)"),
          subtitle = paste0("Effect: ", effect_label, 
                          " (LFC = ", round(p_effect$Log_Fold_Change, 3), 
                          ", q = ", format(p_effect$Q_value, scientific = TRUE, digits = 2), ")"),
          x = "Site (ordered by depth)",
          y = "Relative Abundance"
        )
    }
    
    # Add common theme
    p1 <- p1 +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11),
        panel.grid.minor = element_blank(),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
      ) +
      ylim(0, plot_max_y)
    
    phylum_plots[[p]] <- p1
  }
  
  # Skip if no plots were created
  if(length(phylum_plots) == 0) {
    message("No plots could be created for any of the specified phyla.")
    return(NULL)
  }
  
  # Arrange the plots in a grid
  n_cols <- min(3, length(phylum_plots))  # Max 3 columns
  n_rows <- ceiling(length(phylum_plots) / n_cols)
  
  # Combine plots into a grid
  grid_plot <- gridExtra::grid.arrange(
    grobs = phylum_plots,
    ncol = n_cols,
    top = grid::textGrob(
      paste0("Environmental Effects: ", env_variable, " (", fraction_name, " Fraction)"),
      gp = grid::gpar(fontsize = 16, fontface = "bold")
    )
  )
  
  return(list(grid_plot = grid_plot, individual_plots = phylum_plots))
}

# Function to create side-by-side comparison of motile vs sessile
create_fraction_comparison_plot <- function(motile_results, sessile_results, env_variable = "Turbidity") {
  
  # Get overlapping phyla
  motile_phyla <- motile_results %>% 
    filter(Environmental_Variable == env_variable) %>% 
    pull(Taxon)
  
  sessile_phyla <- sessile_results %>% 
    filter(Environmental_Variable == env_variable) %>% 
    pull(Taxon)
  
  common_phyla <- intersect(motile_phyla, sessile_phyla)
  
  if(length(common_phyla) == 0) {
    message(paste("No common phyla found for", env_variable, "effects"))
    return(NULL)
  }

  # Create comparison data
  comparison_data <- rbind(
    motile_results %>% 
      filter(Environmental_Variable == env_variable & Taxon %in% common_phyla) %>%
      dplyr::mutate(Fraction_Type = "Motile"),
    sessile_results %>% 
      filter(Environmental_Variable == env_variable & Taxon %in% common_phyla) %>%
      dplyr::mutate(Fraction_Type = "Sessile")
  )
  
  # Create comparison plot with improved colors
  comparison_plot <- ggplot(comparison_data, aes(x = Taxon, y = Log_Fold_Change, fill = Fraction_Type)) +
    geom_col(position = "dodge", color = "black", size = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
    scale_fill_manual(
      values = c("Motile" = "#3B8C6E", "Sessile" = "#FEBC2C"),
      name = "Fraction"
    ) +
    labs(
      title = paste0("Comparison of ", env_variable, " Effects: Motile vs Sessile Fractions"),
      x = "Phylum",
      y = "Log Fold Change",
      fill = "Fraction"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold"),
      legend.position = "bottom"
    )
  
  return(comparison_plot)
}

# =============================================================================
# CREATE SIGNIFICANT RESULTS DATA FROM AVAILABLE OBJECTS
# =============================================================================

cat("Creating environmental significant results data...\n")

# Function to extract significant results from environmental results objects
extract_env_significant <- function(env_results_obj, fraction_name, tax_level) {
  
  if(!exists(deparse(substitute(env_results_obj)))) {
    return(data.frame())
  }
  
  results <- get(deparse(substitute(env_results_obj)))
  
  # Extract significant depth effects
  depth_sig <- which(results$diff_depth == TRUE)
  turbidity_sig <- which(results$diff_turbidity_std_rank == TRUE)
  
  # Create depth results
  depth_data <- data.frame()
  if(length(depth_sig) > 0) {
    depth_data <- data.frame(
      Analysis = "Environmental_Effects",
      Fraction = fraction_name,
      Taxonomic_Level = tax_level,
      Environmental_Variable = "Depth",
      Taxon = results$taxon[depth_sig],
      Log_Fold_Change = round(results$lfc_depth[depth_sig], 4),
      P_value = results$p_depth[depth_sig],
      Q_value = results$q_depth[depth_sig],
      Robust = results$passed_ss_depth[depth_sig],
      Direction = ifelse(results$lfc_depth[depth_sig] > 0, "Increases_with_depth", "Decreases_with_depth"),
      stringsAsFactors = FALSE
    )
  }
  
  # Create turbidity results
  turbidity_data <- data.frame()
  if(length(turbidity_sig) > 0) {
    turbidity_data <- data.frame(
      Analysis = "Environmental_Effects",
      Fraction = fraction_name,
      Taxonomic_Level = tax_level,
      Environmental_Variable = "Turbidity",
      Taxon = results$taxon[turbidity_sig],
      Log_Fold_Change = round(results$lfc_turbidity_std_rank[turbidity_sig], 4),
      P_value = results$p_turbidity_std_rank[turbidity_sig],
      Q_value = results$q_turbidity_std_rank[turbidity_sig],
      Robust = results$passed_ss_turbidity_std_rank[turbidity_sig],
      Direction = ifelse(results$lfc_turbidity_std_rank[turbidity_sig] > 0, "Increases_with_turbidity", "Decreases_with_turbidity"),
      stringsAsFactors = FALSE
    )
  }
  
  return(rbind(depth_data, turbidity_data))
}

# Extract significant results from available objects
sig_env_phylum_motile <- data.frame()
sig_env_phylum_sessile <- data.frame()

if(exists("motile_PHY_env_results")) {
  sig_env_phylum_motile <- extract_env_significant(motile_PHY_env_results, "Motile", "Phylum")
  cat("Extracted", nrow(sig_env_phylum_motile), "significant results for motile phyla\n")
}

if(exists("sessile_PHY_env_results")) {
  sig_env_phylum_sessile <- extract_env_significant(sessile_PHY_env_results, "Sessile", "Phylum")
  cat("Extracted", nrow(sig_env_phylum_sessile), "significant results for sessile phyla\n")
}

# Try alternative object names if the above don't exist
if(nrow(sig_env_phylum_motile) == 0 && exists("env_results_phylum_motile")) {
  sig_env_phylum_motile <- extract_env_significant(env_results_phylum_motile, "Motile", "Phylum")
  cat("Extracted", nrow(sig_env_phylum_motile), "significant results for motile phyla (alternative)\n")
}

if(nrow(sig_env_phylum_sessile) == 0 && exists("env_results_phylum_sessile")) {
  sig_env_phylum_sessile <- extract_env_significant(env_results_phylum_sessile, "Sessile", "Phylum")
  cat("Extracted", nrow(sig_env_phylum_sessile), "significant results for sessile phyla (alternative)\n")
}

# =============================================================================
# CREATE PLOTS FOR MOTILE FRACTION
# =============================================================================

cat("Creating environmental effect plots for motile fraction...\n")

# Turbidity effects - Motile
if(nrow(sig_env_phylum_motile) > 0) {
  
  motile_turbidity_plots <- create_environmental_phylum_barplots(
    physeq_obj = motile_physeq,
    significant_results = sig_env_phylum_motile,
    env_variable = "Turbidity",
    fraction_name = "Motile",
    y_margin = 0.2
  )
  
  if(!is.null(motile_turbidity_plots)) {
    ggsave("Motile_Turbidity_Phyla_Cividis.pdf", 
           motile_turbidity_plots$grid_plot,
           width = 16, height = 12, dpi = 300)
    cat("Saved: Motile_Turbidity_Phyla_Cividis.pdf\n")
  }
}

# Depth effects - Motile  
if(nrow(sig_env_phylum_motile) > 0) {
  
  motile_depth_plots <- create_environmental_phylum_barplots(
    physeq_obj = motile_physeq,
    significant_results = sig_env_phylum_motile,
    env_variable = "Depth",
    fraction_name = "Motile",
    y_margin = 0.2
  )
  
  if(!is.null(motile_depth_plots)) {
    ggsave("Motile_Depth_Phyla_Mako.pdf", 
           motile_depth_plots$grid_plot,
           width = 16, height = 12, dpi = 300)
    cat("Saved: Motile_Depth_Phyla_Mako.pdf\n")
  }
}

# =============================================================================
# CREATE PLOTS FOR SESSILE FRACTION
# =============================================================================

cat("Creating environmental effect plots for sessile fraction...\n")

# Turbidity effects - Sessile
if(nrow(sig_env_phylum_sessile) > 0) {
  
  sessile_turbidity_plots <- create_environmental_phylum_barplots(
    physeq_obj = sessile_physeq,
    significant_results = sig_env_phylum_sessile,
    env_variable = "Turbidity",
    fraction_name = "Sessile",
    y_margin = 0.2
  )
  
  if(!is.null(sessile_turbidity_plots)) {
    ggsave("Sessile_Turbidity_Phyla_Cividis.pdf", 
           sessile_turbidity_plots$grid_plot,
           width = 16, height = 12, dpi = 300)
    cat("Saved: Sessile_Turbidity_Phyla_Cividis.pdf\n")
  }
}

# Depth effects - Sessile
if(nrow(sig_env_phylum_sessile) > 0) {
  
  sessile_depth_plots <- create_environmental_phylum_barplots(
    physeq_obj = sessile_physeq,
    significant_results = sig_env_phylum_sessile,
    env_variable = "Depth",
    fraction_name = "Sessile",
    y_margin = 0.2
  )
  
  if(!is.null(sessile_depth_plots)) {
    ggsave("Sessile_Depth_Phyla_Mako.pdf", 
           sessile_depth_plots$grid_plot,
           width = 14, height = 6, dpi = 300)
    cat("Saved: Sessile_Depth_Phyla_Mako.pdf\n")
  }
}

# =============================================================================
# CREATE COMPARISON PLOTS
# =============================================================================

cat("Creating fraction comparison plots...\n")

# Create comparison plots if both motile and sessile data exist
if(nrow(sig_env_phylum_motile) > 0 && nrow(sig_env_phylum_sessile) > 0) {
  
  # Turbidity comparison
  turbidity_comparison <- create_fraction_comparison_plot(
    sig_env_phylum_motile, sig_env_phylum_sessile, "Turbidity"
  )
  
  if(!is.null(turbidity_comparison)) {
    ggsave("Fraction_Comparison_Turbidity_Effects_Final.pdf", turbidity_comparison,
           width = 12, height = 8, dpi = 300)
    cat("Saved: Fraction_Comparison_Turbidity_Effects_Final.pdf\n")
  }
  
  # Depth comparison
  depth_comparison <- create_fraction_comparison_plot(
    sig_env_phylum_motile, sig_env_phylum_sessile, "Depth"
  )
  
  if(!is.null(depth_comparison)) {
    ggsave("Fraction_Comparison_Depth_Effects_Final.pdf", depth_comparison,
           width = 12, height = 8, dpi = 300)
    cat("Saved: Fraction_Comparison_Depth_Effects_Final.pdf\n")
  }
}

# =============================================================================
# SUMMARY
# =============================================================================

cat("\n=== ENVIRONMENTAL PLOTTING SUMMARY ===\n")
cat("Motile significant phyla:", nrow(sig_env_phylum_motile), "\n")
cat("Sessile significant phyla:", nrow(sig_env_phylum_sessile), "\n")

if(nrow(sig_env_phylum_motile) > 0) {
  cat("Motile phyla with effects:\n")
  print(sig_env_phylum_motile %>% select(Taxon, Environmental_Variable, Log_Fold_Change, Q_value, Robust))
}

if(nrow(sig_env_phylum_sessile) > 0) {
  cat("Sessile phyla with effects:\n")
  print(sig_env_phylum_sessile %>% select(Taxon, Environmental_Variable, Log_Fold_Change, Q_value, Robust))
}

cat("\nEnvironmental effects plotting completed!\n")
cat("Check your working directory for PDF files.\n")
```


# LFC take 2 12/4 simpler code - R should not crash
#part 1
```{r}
# RECOVERY SCRIPT - Reload essential 18S ANCOMBC objects from CSV files
# This is much safer than running the full analysis again
# STEP 1: Reload the significant results
cat("Loading significant results from CSV files...\n")

all_environmental_significant <- read.csv("all_environmental_significant_results_18S.csv")
cat("✓ Loaded all_environmental_significant:", nrow(all_environmental_significant), "rows\n")

# STEP 2: Load phyloseq object if saved
if(file.exists("ancombc_raw_counts_physeq_18S.RData")) {
  cat("Loading phyloseq object...\n")
  load("ancombc_raw_counts_physeq_18S.RData")
  cat("✓ Loaded phyloseq object\n")
} else {
  cat("⚠ Phyloseq RData file not found\n")
}

# STEP 3: Create the turbidity_data_18S object (the one we need for plotting)
cat("\nCreating turbidity_data_18S from all_environmental_significant...\n")

# Get taxonomy table for phylum mapping
if(exists("ancom_physeq")) {
  tax_table_df <- as.data.frame(tax_table(ancom_physeq))
  
  family_phylum_map <- tax_table_df %>%
    filter(!is.na(family) & family != "") %>%
    distinct(family, phylum) %>%
    mutate(
      family_clean = gsub("^f__", "", family),
      family_clean = gsub(";.*", "", family_clean),
      phylum_clean = gsub("^p__", "", phylum),
      phylum_clean = gsub(";.*", "", phylum_clean)
    ) %>%
    select(family_clean, phylum_clean)
  
  # Create turbidity_data_18S
  turbidity_data_18S <- all_environmental_significant %>%
    filter(Taxonomic_Level == "Family" & Environmental_Variable == "Turbidity") %>%
    mutate(Family_clean = Taxon) %>%
    left_join(family_phylum_map, by = c("Family_clean" = "family_clean")) %>%
    rename(Phylum = phylum_clean) %>%
    filter(!is.na(Phylum)) %>%
    mutate(Direction_simple = ifelse(Log_Fold_Change > 0, "Increases", "Decreases"))
  
  cat("✓ Created turbidity_data_18S:", nrow(turbidity_data_18S), "families\n")
  cat("  Phyla:", length(unique(turbidity_data_18S$Phylum)), "\n")
  cat("  Robust effects:", sum(turbidity_data_18S$Robust), "\n")
  
} else {
  cat("✗ Cannot create turbidity_data_18S without phyloseq object\n")
}

cat("\n=== RECOVERY COMPLETE ===\n")
cat("Objects now in environment:\n")
cat("- all_environmental_significant\n")
if(exists("ancom_physeq")) cat("- ancom_physeq\n")
if(exists("turbidity_data_18S")) cat("- turbidity_data_18S\n")
```

===============================
----- FUNCTIONAL ANALYSIS ----
==============================

## Functional Trait Analysis: Testing Feeding Strategy as Predictor of Turbidity Sensitivity

Overview
This section tests whether nutritional strategy predicts family-level turbidity responses using 18S ARMS metabarcoding data. The analysis assigns families to functional groups based on feeding mechanisms and tests for directional patterns in turbidity responses (ANCOMBC2 results).

Key Question
Do suspension feeding mechanisms predict sediment tolerance? Specifically, we test whether:
- Fine Dead-End Microfilterers (FDM; sponges) decrease with turbidity (low sediment tolerance)
- Active Dead-End Sieves (ADS; bivalves, barnacles) increase with turbidity (high sediment tolerance via particle rejection)

Required Input Data
1. **ANCOMBC2 results** from three fractions (All_Samples, Motile, Sessile):
   - `env_results_family_all`
   - `env_results_family_motile`
   - `env_results_family_sessile`
   
2. **Functional trait data** (`trait_data_v4`):
   - Must contain columns: `accepted_family`, `nutritional_strategy`, `phyllum`
   - Nutritional strategies use standardized abbreviations (e.g., FDM, ADS, WCP)

3. **Output directory** (optional): Set `output_dir` for saving results, defaults to current directory

What This Code Does
1. **Data preparation**: Cleans trait data, expands abbreviations, merges with ANCOMBC2 results
2. **Functional classification**: Creates binary indicators for 6 suspension feeder mechanisms + other feeding strategies
3. **Statistical analysis**: Calculates mean log-fold changes by functional group, tests against zero
4. **Visualization**: Generates figures with custom color schemes
5. **Export**: Saves summary tables (CSV) and figures (PDF)

Important Notes
- Analysis focuses on **All_Samples fraction** for main results (concordant with COI dataset)
- Functional groups with n<2 families are excluded from statistical tests (insufficient power)
- One photoautotroph outlier (Dumontiaceae, LFC=+1.09) is robust but idiosyncratic
- Large error bars reflect genuine ecological heterogeneity within functional groups, not statistical weakness

Output Files
- `functional_group_summary.csv` - Statistics for all functional groups
- `suspension_mechanism_summary.csv` - Detailed suspension feeder results
- `photoautotroph_subgroups.csv` - Coralline vs. non-coralline patterns
- `functional_groups_all.pdf` - Main figure showing all groups
- `suspension_mechanisms.pdf` - Detailed suspension feeder figure
- `fdm_vs_ads.pdf` - Key prediction test visualization
- `functional_analysis_combined.pdf` - Combined multipanel figure
```{r}
# ============ 1. LOAD AND PREPARE DATA ============

# Load trait data 
#trait_data_v4 <- read.csv("/Users/nicolepittoors/Documents/Documents - biscaya/Bioinformatics/ARMS_18S_LULU/FINAL_18S_Pipeline/trait_mapping_QC_v4.csv", stringsAsFactors = FALSE)

# Set output directory
#if(!exists("output_dir")) output_dir <- "/Users/nicolepittoors/Documents/Documents - biscaya/Bioinformatics/ARMS_18S_LULU/FINAL_18S_Pipeline/Taxa_Output_18S"


#library(dplyr)
#library(tidyr)
#library(stringr)
#library(ggplot2)
#library(patchwork)

# ============ PART 1: DATA PREPARATION ============

# 1.1: Clean trait data and expand abbreviations
abbr_to_full <- c(
  "DEP" = "Deposit_Feeder", "PHOT" = "Photoautotroph",
  "FDM" = "Fine_Dead_End_Microfilterer", "PRED" = "Predator",
  "SCAV" = "Scavenger", "ADS" = "Active_Dead-End_Sieve",
  "WCP" = "Weak_Ciliary_Pumper", "GRAZ" = "Grazer",
  "TPC" = "Truly_Passive_Collectors", "RNM" = "Renewable_Mucus_Net",
  "SYBP" = "Symbiotic_Photosynthesis", "RCF" = "Ram/Crossflow"
)

full_to_abbr <- c(
  "Deposit_Feeder" = "DEP", "Photoautotroph" = "PHOT",
  "Fine_Dead_End_Microfilterer" = "FDM", "Predator" = "PRED",
  "Scavenger" = "SCAV", "Active_Dead-End_Sieve" = "ADS",
  "Weak_Ciliary_Pumper" = "WCP", "Grazer" = "GRAZ",
  "Truly_Passive_Collectors" = "TPC", "Renewable_Mucus_Net" = "RNM",
  "Symbiotic_Photosynthesis" = "SYBP", "Ram/Crossflow" = "RCF",
  "Deposit_feeder" = "DEP", "Parasite" = "PARA",
  "Chemolithoautototroph" = "CHEM", "Unknown" = "UNK"
)

# Expand abbreviations
expand_strategy <- function(x) {
  if(is.na(x) || x == "" || x == "Unknown") return("Unknown")
  parts <- unlist(str_split(x, ";\\s*"))
  expanded <- sapply(parts, function(p) {
    if(p %in% names(abbr_to_full)) return(abbr_to_full[p])
    gsub("_feeder", "_Feeder", gsub("_photosynthesis", "_Photosynthesis", p))
  })
  paste(expanded, collapse = ";")
}

create_abbr <- function(x) {
  if(is.na(x) || x == "" || x == "Unknown") return("UNK")
  parts <- unlist(str_split(x, ";\\s*"))
  abbrs <- sapply(parts, function(p) {
    if(p %in% names(full_to_abbr)) return(full_to_abbr[p])
    return(toupper(substr(gsub("[^A-Za-z]", "", p), 1, 4)))
  })
  if(length(abbrs) > 1) paste(abbrs, collapse = "+") else abbrs[1]
}

# Add family_clean if missing
if(!"family_clean" %in% names(trait_data_v4)) {
  trait_data_v4 <- trait_data_v4 %>%
    mutate(family_clean = tolower(trimws(accepted_family)))
}

# Apply cleaning
trait_data_v4_cleaned <- trait_data_v4 %>%
  mutate(
    nutritional_strategy_full = sapply(nutritional_strategy, expand_strategy),
    nutritional_strategy_abbr = sapply(nutritional_strategy_full, create_abbr),
    primary_function = sapply(nutritional_strategy_full, function(x) {
      str_trim(unlist(str_split(x, ";"))[1])
    }),
    primary_function_abbr = sapply(primary_function, function(x) {
      if(x %in% names(full_to_abbr)) full_to_abbr[x] else "UNK"
    })
  )

# 1.2: Load and combine ANCOM-BC2 results from all fractions
env_results_all <- env_results_family_all %>%
  mutate(
    Fraction = "All_Samples",
    diff_turbidity_logical = as.logical(diff_turbidity_std_rank),
    passed_ss_logical = as.logical(passed_ss_turbidity_std_rank)
  )

env_results_motile_df <- env_results_family_motile %>%
  mutate(
    Fraction = "Motile",
    diff_turbidity_logical = as.logical(diff_turbidity_std_rank),
    passed_ss_logical = as.logical(passed_ss_turbidity_std_rank)
  )

env_results_sessile_df <- env_results_family_sessile %>%
  mutate(
    Fraction = "Sessile", 
    diff_turbidity_logical = as.logical(diff_turbidity_std_rank),
    passed_ss_logical = as.logical(passed_ss_turbidity_std_rank)
  )

full_ancom_results <- rbind(
  env_results_all, env_results_motile_df, env_results_sessile_df
) %>%
  dplyr::select(
    Taxon = taxon, Fraction,
    Log_Fold_Change = lfc_turbidity_std_rank,
    P_value = p_turbidity_std_rank,
    Q_value = q_turbidity_std_rank,
    Is_Significant = diff_turbidity_logical,
    Robust = passed_ss_logical
  ) %>%
  dplyr::mutate(
    Is_Significant = ifelse(is.na(Is_Significant), FALSE, Is_Significant),
    Robust = ifelse(is.na(Robust), FALSE, Robust),
    Direction = case_when(
      Log_Fold_Change > 0 ~ "Increases_with_turbidity",
      Log_Fold_Change < 0 ~ "Decreases_with_turbidity",
      TRUE ~ "No_change"
    ),
    family_clean = tolower(trimws(Taxon))
  )

# 1.3: Merge with trait data and add phylum
# Remove duplicates from trait data (keep first occurrence)
trait_data_v4_cleaned_unique <- trait_data_v4_cleaned %>%
  dplyr::distinct(family_clean, .keep_all = TRUE)

functional_all_families <- full_ancom_results %>%
  dplyr::left_join(trait_data_v4_cleaned_unique, by = "family_clean") %>%
  dplyr::mutate(
    final_phylum = coalesce(phyllum, case_when(
      str_detect(tolower(Taxon), "rhodophyta") ~ "Rhodophyta",
      str_detect(tolower(Taxon), "porifera") ~ "Porifera",
      str_detect(tolower(Taxon), "bryozoa") ~ "Bryozoa",
      str_detect(tolower(Taxon), "cnidaria") ~ "Cnidaria",
      str_detect(tolower(Taxon), "annelida") ~ "Annelida",
      str_detect(tolower(Taxon), "arthropoda") ~ "Arthropoda",
      str_detect(tolower(Taxon), "mollusca") ~ "Mollusca",
      str_detect(tolower(Taxon), "chordata") ~ "Chordata",
      TRUE ~ "Other"
    ))
  )

# ============ PART 2: CREATE FUNCTIONAL INDICATORS ============

functional_dataset <- functional_all_families %>%
  mutate(
    # Suspension feeders (6 mechanistic categories)
    is_FDM = case_when(
      str_detect(nutritional_strategy_full, "Fine_Dead_End_Microfilterer") ~ 1,
      final_phylum == "Porifera" ~ 1, TRUE ~ 0
    ),
    is_ADS = case_when(
      str_detect(nutritional_strategy_full, "Active_Dead-End_Sieve") ~ 1, TRUE ~ 0
    ),
    is_WCP = case_when(
      str_detect(nutritional_strategy_full, "Weak_Ciliary_Pumper") ~ 1,
      final_phylum == "Bryozoa" ~ 1, TRUE ~ 0
    ),
    is_TPC = case_when(
      str_detect(nutritional_strategy_full, "Truly_Passive_Collectors") ~ 1, TRUE ~ 0
    ),
    is_RNM = case_when(
      str_detect(nutritional_strategy_full, "Renewable_Mucus_Net") ~ 1, TRUE ~ 0
    ),
    is_RCF = case_when(
      str_detect(nutritional_strategy_full, "Ram/Crossflow") ~ 1, TRUE ~ 0
    ),
    is_any_suspension = case_when(
      is_FDM == 1 | is_ADS == 1 | is_WCP == 1 | 
      is_TPC == 1 | is_RNM == 1 | is_RCF == 1 ~ 1, TRUE ~ 0
    ),
    
    # Other feeding strategies
    is_deposit_feeder = case_when(
      str_detect(nutritional_strategy_full, "Deposit_Feeder") ~ 1, TRUE ~ 0
    ),
    is_predator = case_when(
      str_detect(nutritional_strategy_full, "Predator") ~ 1, TRUE ~ 0
    ),
    is_photoautotroph = case_when(
      str_detect(nutritional_strategy_full, "Photoautotroph") ~ 1,
      final_phylum == "Rhodophyta" ~ 1, TRUE ~ 0
    ),
    is_grazer = case_when(
      str_detect(nutritional_strategy_full, "Grazer") ~ 1, TRUE ~ 0
    ),
    is_scavenger = case_when(
      str_detect(nutritional_strategy_full, "Scavenger") ~ 1, TRUE ~ 0
    ),
    is_parasite = case_when(
      str_detect(nutritional_strategy_full, "Parasite") ~ 1, TRUE ~ 0
    )
  )

# ============ PART 3: STATISTICAL ANALYSIS ============

# 3.1: Summary statistics by functional group
functional_group_summary <- functional_dataset %>%
  filter(as.logical(Is_Significant), Fraction == "All_Samples") %>%
  group_by(primary_function_abbr) %>%
  summarise(
    n_families = n(),
    mean_lfc = mean(Log_Fold_Change),
    median_lfc = median(Log_Fold_Change),
    sd_lfc = sd(Log_Fold_Change),
    se_lfc = sd_lfc / sqrt(n_families),
    n_increase = sum(Direction == "Increases_with_turbidity"),
    n_decrease = sum(Direction == "Decreases_with_turbidity"),
    percent_decrease = round(n_decrease / n_families * 100, 1),
    # One-sample t-test against zero
    t_stat = ifelse(n_families > 1, mean_lfc / se_lfc, NA),
    p_value = ifelse(n_families > 1, 
                     2 * pt(-abs(t_stat), df = n_families - 1), NA),
    .groups = "drop"
  ) %>%
  filter(!is.na(primary_function_abbr), primary_function_abbr != "UNK", n_families >= 2) %>%
  arrange(desc(mean_lfc))

# 3.2: Suspension feeder detailed analysis
suspension_summary <- functional_dataset %>%
  dplyr::filter(as.logical(Is_Significant), Fraction == "All_Samples") %>%
  dplyr::select(Taxon, Log_Fold_Change, Direction, final_phylum,
         is_FDM, is_ADS, is_WCP, is_TPC, is_RNM, is_RCF) %>%
  pivot_longer(cols = c(is_FDM, is_ADS, is_WCP, is_TPC, is_RNM, is_RCF),
               names_to = "suspension_type", values_to = "has_type") %>%
  dplyr::filter(has_type == 1) %>%
  dplyr::mutate(
    Suspension_Abbr = case_when(
      suspension_type == "is_FDM" ~ "FDM",
      suspension_type == "is_ADS" ~ "ADS",
      suspension_type == "is_WCP" ~ "WCP",
      suspension_type == "is_TPC" ~ "TPC",
      suspension_type == "is_RNM" ~ "RNM",
      suspension_type == "is_RCF" ~ "RCF"
    )
  ) %>%
  group_by(Suspension_Abbr) %>%
  summarise(
    n = n(),
    mean_lfc = mean(Log_Fold_Change),
    se_lfc = sd(Log_Fold_Change) / sqrt(n),
    n_decrease = sum(Direction == "Decreases_with_turbidity"),
    percent_decrease = round(n_decrease / n * 100, 1),
    .groups = "drop"
  )

# 3.3: Photoautotroph subgroup analysis
photoautotroph_analysis <- functional_dataset %>%
  dplyr::filter(is_photoautotroph == 1, as.logical(Is_Significant), 
         Fraction == "All_Samples") %>%
  dplyr::mutate(
    is_coralline = Taxon %in% c("Hapalidiaceae", "Mesophyllumaceae", 
                                "Lithophyllaceae", "Corallinaceae")
  ) %>%
  group_by(is_coralline) %>%
  summarise(
    n = n(),
    mean_lfc = mean(Log_Fold_Change),
    se_lfc = sd(Log_Fold_Change) / sqrt(n),
    .groups = "drop"
  )

# ============ PART 4: VISUALIZATION ============

# Color schemes
phylum_color_map <- c(
  "Annelida" = "#60714E", "Arthropoda" = "#3B8C6E", 
  "Bryozoa" = "#5BBFB0", "Chordata" = "#408BA8",
  "Cnidaria" = "#594A9E", "Echinodermata" = "#A8B1D6",
  "Mollusca" = "#F69B9C", "Rhodophyta" = "#F16363",
  "Porifera" = "#F68C30", "Unknown" = "grey40"
)

suspension_colors <- c(
  "FDM" = "#F68C30", "ADS" = "#3B8C6E", "WCP" = "#5BBFB0",
  "TPC" = "#A8B1D6", "RNM" = "#408BA8", "RCF" = "#408BA8"
)

direction_colors <- c(
  "Increases_with_turbidity" = "#3B8C6E",
  "Decreases_with_turbidity" = "#F68C30"
)

# 4.1: Main functional groups plot
sig_functional_all <- functional_dataset %>%
  dplyr::filter(as.logical(Is_Significant), Fraction == "All_Samples")

plot_functional_groups <- functional_group_summary %>%
  mutate(
    Group_Type = ifelse(primary_function_abbr %in% c("FDM", "ADS", "WCP", "TPC", "RNM", "RCF"),
                       "Suspension Feeder", "Other")
  ) %>%
  ggplot(aes(x = reorder(primary_function_abbr, mean_lfc), 
             y = mean_lfc, fill = Group_Type)) +
  geom_col(width = 0.7, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = mean_lfc - se_lfc, 
                    ymax = mean_lfc + se_lfc),
                width = 0.3, size = 0.5) +
  geom_text(aes(label = paste0("n=", n_families)), 
            hjust = ifelse(functional_group_summary$mean_lfc > 0, -0.3, 1.3),
            size = 3.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  scale_fill_manual(values = c("Suspension Feeder" = "#56B4E9", "Other" = "#E69F00")) +
  labs(title = "Functional Group Responses to Turbidity",
       x = "Functional Group", y = "Mean Log-Fold Change",
       fill = "Group Type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

# 4.2: Suspension feeders detailed
suspension_long <- functional_dataset %>%
  dplyr::filter(as.logical(Is_Significant), Fraction == "All_Samples") %>%
  dplyr::select(Taxon, Log_Fold_Change, Direction, is_FDM, is_ADS, is_WCP, 
         is_TPC, is_RNM, is_RCF) %>%
  pivot_longer(cols = c(is_FDM, is_ADS, is_WCP, is_TPC, is_RNM, is_RCF),
               names_to = "type", values_to = "has") %>%
  dplyr::filter(has == 1) %>%
  dplyr::mutate(Category = gsub("is_", "", type))

plot_suspension <- suspension_long %>%
  left_join(suspension_summary, by = c("Category" = "Suspension_Abbr")) %>%
  ggplot(aes(x = reorder(Category, mean_lfc), y = mean_lfc, fill = Category)) +
  geom_col(width = 0.7, color = "black") +
  geom_errorbar(aes(ymin = mean_lfc - se_lfc, ymax = mean_lfc + se_lfc),
                width = 0.3) +
  geom_text(aes(label = paste0("n=", n)), vjust = -0.5, size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  scale_fill_manual(values = suspension_colors) +
  labs(title = "Suspension Feeder Mechanisms",
       x = NULL, y = "Mean Log-Fold Change") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# 4.3: FDM vs ADS comparison
plot_fdm_ads <- suspension_long %>%
  dplyr::filter(Category %in% c("FDM", "ADS")) %>%
  ggplot(aes(x = Category, y = Log_Fold_Change, fill = Category)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(aes(color = Direction), width = 0.2, size = 2.5, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = suspension_colors) +
  scale_color_manual(values = direction_colors) +
  labs(title = "FDM vs ADS: Key Prediction Test",
       x = NULL, y = "Log-Fold Change") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  guides(fill = "none")

# Save plots
ggsave(file.path(output_dir,"functional_groups_all.pdf"), plot_functional_groups, 
       width = 10, height = 8, dpi = 300)
ggsave(file.path(output_dir, "suspension_mechanisms.pdf"), plot_suspension, 
       width = 8, height = 6, dpi = 300)
ggsave(file.path(output_dir,"fdm_vs_ads.pdf"), plot_fdm_ads, 
       width = 6, height = 6, dpi = 300)

# Combined figure
combined <- (plot_suspension / plot_fdm_ads) | plot_functional_groups
ggsave(file.path(output_dir,"functional_analysis_combined.pdf"), combined, 
       width = 14, height = 10, dpi = 300)

# ============ PART 5: EXPORT RESULTS ============

write.csv(functional_group_summary, file.path(output_dir, "functional_group_summary.csv"), row.names = FALSE)
write.csv(suspension_summary, file.path(output_dir, "suspension_mechanism_summary.csv"), row.names = FALSE)
write.csv(photoautotroph_analysis, file.path(output_dir, "photoautotroph_subgroups.csv"), row.names = FALSE)

cat("\n=== ANALYSIS COMPLETE ===\n")
cat("Functional groups analyzed:", nrow(functional_group_summary), "\n")
cat("Suspension mechanisms tested:", nrow(suspension_summary), "\n")
cat("Results exported to CSV files and PDF figures\n")
```


```{r vertical phylum LFC plots, all fractions}

library(ggplot2)
library(dplyr)

cat("=== CREATING CLEAN PHYLUM LFC PLOTS ===\n\n")

# Get phylum-level turbidity effects
phylum_turb <- all_environmental_significant %>%
  filter(Taxonomic_Level == "Phylum" & Environmental_Variable == "Turbidity")

# Custom colors
color_increase <- "#3B8C6E"  # Green for increases
color_decrease <- "#F68C30"  # Orange for decreases

# ALL SAMPLES
all_phyla <- phylum_turb %>%
  filter(Fraction == "All_Samples") %>%
  arrange(Log_Fold_Change) %>%
  mutate(
    Phylum_ordered = factor(Taxon, levels = Taxon),
    Direction = ifelse(Log_Fold_Change > 0, "Increases_with_turbidity", "Decreases_with_turbidity")
  )

if(nrow(all_phyla) > 0) {
  cat("\n=== ALL SAMPLES ===\n")
  cat("Phyla found:", nrow(all_phyla), "\n")
  print(all_phyla %>% select(Taxon, Log_Fold_Change, Q_value, Robust))
  
  p_all <- ggplot(all_phyla, aes(x = Phylum_ordered, y = Log_Fold_Change, fill = Direction)) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.7) +
    scale_fill_manual(
      values = c(
        "Increases_with_turbidity" = color_increase,
        "Decreases_with_turbidity" = color_decrease
      ),
      name = "",
      labels = c("Decreases with turbidity", "Increases with turbidity")
    ) +
    labs(
      title = "All fractions",
      x = "",
      y = "Log₂ Fold Change"
    ) +
    coord_flip() +  # VERTICAL PLOT
    theme_minimal(base_size = 11) +
    theme(
      # All text black
      text = element_text(color = "black"),
      axis.text = element_text(color = "black", size = 10),
      axis.title = element_text(color = "black", size = 11, face = "bold"),
      plot.title = element_text(color = "black", size = 12, face = "bold", hjust = 0.5),
      legend.text = element_text(color = "black", size = 9),
      
      # Remove gridlines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # White background
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      
      # Add axis lines
      axis.line = element_line(color = "black", linewidth = 0.5),
      
      # Legend
      legend.position = "bottom",
      legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
      legend.key.size = unit(0.5, "cm")
    )
  
  print(p_all)
  
  ggsave("Phylum_LFC_All_Clean_18S.pdf", p_all, width = 7, height = 5, dpi = 300)
  ggsave("Phylum_LFC_All_Clean_18S.png", p_all, width = 7, height = 5, dpi = 600)
  
  cat("\n✓ All fractions plot saved\n")
}

# MOTILE
motile_phyla <- phylum_turb %>%
  filter(Fraction == "Motile") %>%
  arrange(Log_Fold_Change) %>%
  mutate(
    Phylum_ordered = factor(Taxon, levels = Taxon),
    Direction = ifelse(Log_Fold_Change > 0, "Increases_with_turbidity", "Decreases_with_turbidity")
  )

if(nrow(motile_phyla) > 0) {
  cat("\n=== MOTILE ===\n")
  cat("Phyla found:", nrow(motile_phyla), "\n")
  print(motile_phyla %>% select(Taxon, Log_Fold_Change, Q_value, Robust))
  
  p_motile <- ggplot(motile_phyla, aes(x = Phylum_ordered, y = Log_Fold_Change, fill = Direction)) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.7) +
    scale_fill_manual(
      values = c(
        "Increases_with_turbidity" = color_increase,
        "Decreases_with_turbidity" = color_decrease
      ),
      name = "",
      labels = c("Decreases with turbidity", "Increases with turbidity")
    ) +
    labs(
      title = "Motile fraction",
      x = "",
      y = "Log₂ Fold Change"
    ) +
    coord_flip() +  # VERTICAL PLOT
    theme_minimal(base_size = 11) +
    theme(
      # All text black
      text = element_text(color = "black"),
      axis.text = element_text(color = "black", size = 10),
      axis.title = element_text(color = "black", size = 11, face = "bold"),
      plot.title = element_text(color = "black", size = 12, face = "bold", hjust = 0.5),
      legend.text = element_text(color = "black", size = 9),
      
      # Remove gridlines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # Add axis lines
      axis.line = element_line(color = "black", linewidth = 0.5),
      
      # Legend
      legend.position = "bottom",
      legend.key.size = unit(0.5, "cm")
    )
  
  print(p_motile)
  
  ggsave("Phylum_LFC_Motile_Clean_18S.pdf", p_motile, width = 7, height = 5, dpi = 300)
  ggsave("Phylum_LFC_Motile_Clean_18S.png", p_motile, width = 7, height = 5, dpi = 600)
  
  cat("\n✓ Motile plot saved\n")
}

# SESSILE
sessile_phyla <- phylum_turb %>%
  filter(Fraction == "Sessile") %>%
  arrange(Log_Fold_Change) %>%
  mutate(
    Phylum_ordered = factor(Taxon, levels = Taxon),
    Direction = ifelse(Log_Fold_Change > 0, "Increases_with_turbidity", "Decreases_with_turbidity")
  )

if(nrow(sessile_phyla) > 0) {
  cat("\n=== SESSILE ===\n")
  cat("Phyla found:", nrow(sessile_phyla), "\n")
  print(sessile_phyla %>% select(Taxon, Log_Fold_Change, Q_value, Robust))
  
  p_sessile <- ggplot(sessile_phyla, aes(x = Phylum_ordered, y = Log_Fold_Change, fill = Direction)) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.7) +
    scale_fill_manual(
      values = c(
        "Increases_with_turbidity" = color_increase,
        "Decreases_with_turbidity" = color_decrease
      ),
      name = "",
      labels = c("Decreases with turbidity", "Increases with turbidity")
    ) +
    labs(
      title = "Sessile fraction",
      x = "",
      y = "Log₂ Fold Change"
    ) +
    coord_flip() +  # VERTICAL PLOT
    theme_minimal(base_size = 11) +
    theme(
      # All text black
      text = element_text(color = "black"),
      axis.text = element_text(color = "black", size = 10),
      axis.title = element_text(color = "black", size = 11, face = "bold"),
      plot.title = element_text(color = "black", size = 12, face = "bold", hjust = 0.5),
      legend.text = element_text(color = "black", size = 9),
      
      # Remove gridlines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # White background
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      
      # Add axis lines
      axis.line = element_line(color = "black", linewidth = 0.5),
      
      # Legend
      legend.position = "bottom",
      legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
      legend.key.size = unit(0.5, "cm")
    )
  
  print(p_sessile)
  
  ggsave("Phylum_LFC_Sessile_Clean_18S.pdf", p_sessile, width = 7, height = 5, dpi = 300)
  ggsave("Phylum_LFC_Sessile_Clean_18S.png", p_sessile, width = 7, height = 5, dpi = 600)
  
  cat("\n✓ Sessile plot saved\n")
}

# SUMMARY
cat("\n==========================================\n")
cat("CLEAN PHYLUM LFC PLOTS COMPLETE!\n")
cat("==========================================\n\n")

cat("Files created:\n")
if(exists("p_all")) cat("1. Phylum_LFC_All_Clean_18S.pdf (All fractions)\n")
if(exists("p_motile")) cat("2. Phylum_LFC_Motile_Clean_18S.pdf (Motile)\n")
if(exists("p_sessile")) cat("3. Phylum_LFC_Sessile_Clean_18S.pdf (Sessile)\n")
```

# COMBINED 18S + COI PHYLUM LFC PLOTS
Creates vertical plots with slashed lines for COI bars
Run this in your 18S R window AFTER exporting COI data
```{r COMBINED 18S + COI PHYLUM LFC PLOTS}
#library(ggplot2)
#library(dplyr)
#library(ggpattern)  # For patterned bars

cat("=== CREATING COMBINED 18S + COI PLOTS ===\n\n")

# Load 18S phylum data (already in environment)
phylum_turb_18S <- all_environmental_significant %>%
  filter(Taxonomic_Level == "Phylum" & Environmental_Variable == "Turbidity") %>%
  mutate(Dataset = "18S rRNA")

# Load COI phylum data (exported from COI window)
phylum_turb_COI <- read.csv("COI_phylum_turbidity_simple.csv")

cat("18S phyla:", nrow(phylum_turb_18S), "\n")
cat("COI phyla:", nrow(phylum_turb_COI), "\n\n")

# Combine both datasets
combined_phylum <- rbind(
  phylum_turb_18S %>% select(Fraction, Taxon, Log_Fold_Change, Q_value, Robust, Direction, Dataset),
  phylum_turb_COI
)

# Custom colors
color_increase <- "#3B8C6E"  # Green
color_decrease <- "#F68C30"  # Orange

# FUNCTION TO CREATE COMBINED PLOT
create_combined_plot <- function(data, fraction_name) {
  
  # Filter to specific fraction
  frac_data <- data %>%
    filter(Fraction == fraction_name) %>%
    arrange(Dataset, Log_Fold_Change) %>%
    mutate(
      # Create a unique ID for each phylum-dataset combination
      Phylum_Dataset = paste0(Taxon, " (", Dataset, ")"),
      Phylum_Dataset_ordered = factor(Phylum_Dataset, levels = unique(Phylum_Dataset)),
      # Simplify direction names
      Direction_simple = ifelse(grepl("Increase", Direction), "Increases_with_turbidity", "Decreases_with_turbidity"),
      # Pattern: solid for 18S, slashed for COI
      Pattern_type = ifelse(Dataset == "18S rRNA", "none", "stripe")
    )
  
  if(nrow(frac_data) == 0) {
    cat("No data for", fraction_name, "\n")
    return(NULL)
  }
  
  cat("\n", fraction_name, ":\n", sep = "")
  cat("  Total entries:", nrow(frac_data), "\n")
  cat("  18S phyla:", sum(frac_data$Dataset == "18S rRNA"), "\n")
  cat("  COI phyla:", sum(frac_data$Dataset == "COI"), "\n")
  
  # Create plot with ggpattern
  p <- ggplot(frac_data, aes(x = Phylum_Dataset_ordered, y = Log_Fold_Change, 
                             fill = Direction_simple, pattern = Pattern_type)) +
    
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    
    # Use geom_bar_pattern for patterned bars
    ggpattern::geom_bar_pattern(
      stat = "identity",
      color = "black",
      linewidth = 0.3,
      width = 0.7,
      pattern_fill = "black",
      pattern_angle = 45,
      pattern_density = 0.1,
      pattern_spacing = 0.025,
      pattern_key_scale_factor = 0.6
    ) +
    
    scale_pattern_manual(values = c("none" = "none", "stripe" = "stripe"), guide = "none") +
    
    scale_fill_manual(
      values = c(
        "Increases_with_turbidity" = color_increase,
        "Decreases_with_turbidity" = color_decrease
      ),
      name = "",
      labels = c("Decreases with turbidity", "Increases with turbidity")
    ) +
    
    labs(
      title = paste0(fraction_name, " fraction - 18S rRNA + COI"),
      subtitle = "Solid bars = 18S rRNA; Striped bars = COI",
      x = "",
      y = "Log₂ Fold Change"
    ) +
    
    coord_flip() +
    
    theme_minimal(base_size = 11) +
    theme(
      # All text black
      text = element_text(color = "black"),
      axis.text = element_text(color = "black", size = 9),
      axis.title = element_text(color = "black", size = 11, face = "bold"),
      plot.title = element_text(color = "black", size = 12, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(color = "black", size = 9, hjust = 0.5),
      legend.text = element_text(color = "black", size = 9),
      
      # Remove gridlines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # White background
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      
      # Axis lines
      axis.line = element_line(color = "black", linewidth = 0.5),
      
      # Legend
      legend.position = "bottom",
      legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3)
    )
  
  return(list(plot = p, data = frac_data))
}

# Create plots for each fraction
cat("\n=== CREATING PLOTS ===\n")

# ALL SAMPLES
all_result <- create_combined_plot(combined_phylum, "All_Samples")
if(!is.null(all_result)) {
  print(all_result$plot)
  ggsave("Combined_18S_COI_Phylum_All.pdf", all_result$plot, width = 8, height = 7, dpi = 300)
  ggsave("Combined_18S_COI_Phylum_All.png", all_result$plot, width = 8, height = 7, dpi = 600)
  cat("✓ All samples saved\n")
}

# MOTILE
motile_result <- create_combined_plot(combined_phylum, "Motile")
if(!is.null(motile_result)) {
  print(motile_result$plot)
  ggsave("Combined_18S_COI_Phylum_Motile.pdf", motile_result$plot, width = 8, height = 7, dpi = 300)
  ggsave("Combined_18S_COI_Phylum_Motile.png", motile_result$plot, width = 8, height = 7, dpi = 600)
  cat("✓ Motile saved\n")
}

# SESSILE
sessile_result <- create_combined_plot(combined_phylum, "Sessile")
if(!is.null(sessile_result)) {
  print(sessile_result$plot)
  ggsave("Combined_18S_COI_Phylum_Sessile.pdf", sessile_result$plot, width = 8, height = 7, dpi = 300)
  ggsave("Combined_18S_COI_Phylum_Sessile.png", sessile_result$plot, width = 8, height = 7, dpi = 600)
  cat("✓ Sessile saved\n")
}

# Save combined data for reference
write.csv(combined_phylum, "Combined_18S_COI_phylum_data.csv", row.names = FALSE)
```



# --------------- EXPLORATORY POST-HOC ANALYSIS: PHOTOAUTOTROPH SUBGROUPS --------------
```{r photoautotroph exploration}
# Define coralline families (calcified crustose forms)
coralline_families <- c("Hapalidiaceae", "Mesophyllumaceae", "Lithophyllaceae", 
                        "Corallinaceae", "Sporolithaceae")

# Create subgroup classification
photoautotroph_subgroups <- functional_dataset %>%
  filter(is_photoautotroph == 1, as.logical(Is_Significant), 
         Fraction == "All_Samples") %>%
  mutate(
    algal_subgroup = case_when(
      Taxon %in% coralline_families ~ "Coralline_Algae",
      TRUE ~ "Fleshy_Algae"
    )
  )

# Calculate subgroup statistics
subgroup_stats <- photoautotroph_subgroups %>%
  group_by(algal_subgroup) %>%
  summarise(
    n_families = n(),
    mean_lfc = mean(Log_Fold_Change),
    median_lfc = median(Log_Fold_Change),
    sd_lfc = sd(Log_Fold_Change),
    se_lfc = sd_lfc / sqrt(n_families),
    n_increase = sum(Direction == "Increases_with_turbidity"),
    n_decrease = sum(Direction == "Decreases_with_turbidity"),
    percent_decrease = round(n_decrease / n_families * 100, 1),
    .groups = "drop"
  )

cat("=== EXPLORATORY SUBGROUP ANALYSIS ===\n")
print(subgroup_stats)

# Statistical comparison between subgroups
welch_test <- t.test(
  Log_Fold_Change ~ algal_subgroup,
  data = photoautotroph_subgroups
)

cat("\nWelch's t-test (Coralline vs Fleshy):\n")
cat("t =", round(welch_test$statistic, 3), "\n")
cat("df =", round(welch_test$parameter, 2), "\n")
cat("p-value =", format.pval(welch_test$p.value, digits = 3), "\n")
cat("Mean difference =", round(diff(welch_test$estimate), 3), "\n")

# Test each subgroup against zero
coralline_test <- t.test(
  photoautotroph_subgroups %>% 
    filter(algal_subgroup == "Coralline_Algae") %>% 
    pull(Log_Fold_Change)
)

fleshy_test <- t.test(
  photoautotroph_subgroups %>% 
    filter(algal_subgroup == "Fleshy_Algae") %>% 
    pull(Log_Fold_Change)
)

cat("\nOne-sample tests (against zero):\n")
cat("Coralline: t =", round(coralline_test$statistic, 3), 
    ", p =", format.pval(coralline_test$p.value, digits = 3), "\n")
cat("Fleshy: t =", round(fleshy_test$statistic, 3), 
    ", p =", format.pval(fleshy_test$p.value, digits = 3), "\n")

# Create visualization
subgroup_plot <- photoautotroph_subgroups %>%
  ggplot(aes(x = algal_subgroup, y = Log_Fold_Change, fill = algal_subgroup)) +
  geom_violin(alpha = 0.3, trim = TRUE) +
  geom_jitter(aes(color = Direction), width = 0.2, size = 3, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 4, shape = 23, 
               fill = "black", color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, 
               size = 1, color = "black") +
  scale_fill_manual(values = c("Coralline_Algae" = "#F16363", 
                               "Fleshy_Algae" = "#F69B9C")) +
  scale_color_manual(values = c("Increases_with_turbidity" = "#3B8C6E",
                                "Decreases_with_turbidity" = "#F68C30")) +
  labs(
    title = "Exploratory Post-Hoc Analysis: Photoautotroph Subgroups",
    subtitle = paste0("Welch's t-test: p = ", 
                     format.pval(welch_test$p.value, digits = 2)),
    x = NULL,
    y = "Log-Fold Change (Turbidity Effect)",
    color = "Response Direction"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  guides(fill = "none")

print(subgroup_plot)

# Export
ggsave("photoautotroph_subgroup_exploratory.pdf", subgroup_plot, 
       width = 8, height = 6, dpi = 300)
```

# FUNCTIONAL COMPOSITION ANALYSIS - UPDATED WITH MECHANISTIC CATEGORIES
Uses revised 6 suspension feeder mechanisms + other functional groups
```{r functional composition analysis}

#library(dplyr)
#library(ggplot2)
#library(tidyr)
#library(stringr)
#library(phyloseq)
#library(patchwork)

# ============ SETUP: COLOR SCHEMES AND SITE ORDER ============

# Site colors (unchanged)
c12_sites <- c("Alderdice_deep1" = "#0E4077", "Alderdice_deep2" = "#0E4077",
               "Alderdice_shallow" = "#9DC9D3", "Bright_deep1" = "#3B8C6E",
               "Bright_deep2" = "#3B8C6E", "Bright_shallow" = "#B0DBBB",
               "Diaphus_deep2" = "#996599", "Diaphus_deep1" = "#996599",
               "EFGB_deep" = "#F68C30", "EFGB_shallow" = "#FEBC2C",
               "McGrail_deep" = "#015951", "Stetson_shallow" = "#FF595E")

# UPDATED: Functional group colors matching new categories
functional_colors <- c(
  "PHOT" = "#F16363",     # Photoautotrophs (Rhodophyta)
  "FDM" = "#F68C30",      # Fine Dead-End Microfilterers (Porifera)
  "ADS" = "#3B8C6E",      # Active Dead-End Sieves
  "WCP" = "#5BBFB0",      # Weak Ciliary Pumpers (Bryozoa)
  "TPC" = "#A8B1D6",      # Truly Passive Collectors
  "RNM" = "#408BA8",      # Renewable Mucus Nets (Ascidians)
  "RCF" = "#408BA8",      # Ram/Crossflow (fish)
  "DEP" = "#60714E",      # Deposit Feeders
  "PRED" = "#594A9E",     # Predators
  "GRAZ" = "#F69B9C",     # Grazers
  "SCAV" = "#C74238",     # Scavengers
  "PARA" = "#812A5A",     # Parasites
  "UNK" = "grey70"        # Unclassified
)

# Site order (high to low turbidity)
turb_ordered_sites <- c("McGrail_deep", "EFGB_shallow", "Alderdice_shallow", 
                        "Bright_shallow", "Diaphus_deep2", "EFGB_deep", 
                        "Diaphus_deep1", "Bright_deep2", "Bright_deep1", 
                        "Alderdice_deep1", "Stetson_shallow", "Alderdice_deep2")

# ============ 1. USE CLEANED TRAIT DATA FROM PREVIOUS ANALYSIS ============

# This assumes you've already run the main functional analysis
# and have trait_data_v4_cleaned_unique available

# If not, load and clean:
if(!exists("trait_data_v4_cleaned_unique")) {
  source("path_to_trait_cleaning_script.R")
}

# Create primary function using the updated classification
functional_dataset_updated <- trait_data_v4_cleaned_unique %>%
  mutate(
    # Determine primary function based on trait and phylum
    primary_function_abbr = case_when(
      # Photoautotrophs first
      str_detect(nutritional_strategy_full, "Photoautotroph") ~ "PHOT",
      phyllum == "Rhodophyta" ~ "PHOT",  # Changed from final_phylum
      
      # Suspension feeders by mechanism
      str_detect(nutritional_strategy_full, "Fine_Dead_End_Microfilterer") ~ "FDM",
      phyllum == "Porifera" ~ "FDM",  # Changed from final_phylum
      
      str_detect(nutritional_strategy_full, "Active_Dead-End_Sieve") ~ "ADS",
      
      str_detect(nutritional_strategy_full, "Weak_Ciliary_Pumper") ~ "WCP",
      phyllum == "Bryozoa" ~ "WCP",  # Changed from final_phylum
      
      str_detect(nutritional_strategy_full, "Truly_Passive_Collectors") ~ "TPC",
      
      str_detect(nutritional_strategy_full, "Renewable_Mucus_Net") ~ "RNM",
      
      str_detect(nutritional_strategy_full, "Ram/Crossflow") ~ "RCF",
      
      # Other feeding strategies
      str_detect(nutritional_strategy_full, "Deposit_Feeder") ~ "DEP",
      str_detect(nutritional_strategy_full, "Predator") ~ "PRED",
      str_detect(nutritional_strategy_full, "Grazer") ~ "GRAZ",
      str_detect(nutritional_strategy_full, "Scavenger") ~ "SCAV",
      str_detect(nutritional_strategy_full, "Parasite") ~ "PARA",
      
      TRUE ~ "UNK"
    )
  )

# ============ 2. EXTRACT AND PREPARE PHYLOSEQ DATA ============

cat("Extracting data from phyloseq object...\n")

family_site_abundances <- ancom_physeq %>%
  tax_glom(taxrank = "family") %>%
  psmelt() %>%
  mutate(family_clean = tolower(trimws(family)))

cat("Families found:", length(unique(family_site_abundances$family_clean)), "\n")

# ============ 3. JOIN WITH FUNCTIONAL TRAITS ============

functional_site_data <- family_site_abundances %>%
  left_join(
    functional_dataset_updated %>% 
      dplyr::select(family_clean, primary_function_abbr) %>% 
      distinct(),
    by = "family_clean"
  ) %>%
  dplyr::mutate(
    # Replace NA with UNK for unassigned families
    primary_function_abbr = ifelse(is.na(primary_function_abbr), "UNK", primary_function_abbr)
  )

cat("Families with trait assignments:", 
    sum(functional_site_data$primary_function_abbr != "UNK"), "/",
    nrow(functional_site_data), "\n")

# ============ 4. CALCULATE FUNCTIONAL COMPOSITION BY SITE ============

functional_composition <- functional_site_data %>%
  group_by(sitelocality, primary_function_abbr) %>%
  dplyr::summarize(
    turbidity_std_rank = first(turbidity_std_rank),
    group_abundance = sum(Abundance),
    .groups = "drop"
  ) %>%
  group_by(sitelocality) %>%
  dplyr::mutate(
    total_abundance = sum(group_abundance),
    Proportion = group_abundance / total_abundance * 100
  ) %>%
  ungroup() %>%
  dplyr::select(sitelocality, turbidity_std_rank, primary_function_abbr, Proportion) %>%
  pivot_wider(names_from = primary_function_abbr, 
              values_from = Proportion, 
              values_fill = 0) %>%
  rename(Site = sitelocality, Turbidity = turbidity_std_rank)

# Verify proportions sum to 100%
check_totals <- functional_composition %>%
  dplyr::select(-Site, -Turbidity) %>%
  rowSums()

cat("All sites sum to 100%:", all(abs(check_totals - 100) < 0.01), "\n")

# ============ 5. RESHAPE FOR PLOTTING ============

composition_long <- functional_composition %>%
  dplyr::select(-Turbidity) %>%
  pivot_longer(cols = -Site, 
               names_to = "Functional_Group", 
               values_to = "Proportion") %>%
  dplyr::filter(!is.na(Functional_Group), Functional_Group != "NA") %>%
  dplyr::mutate(
    Site = factor(Site, levels = turb_ordered_sites),
    # Order groups: photoautotrophs, suspension feeders, then others
    Functional_Group = factor(Functional_Group, 
                             levels = c("PHOT", "FDM", "ADS", "WCP", "TPC", "RNM", "RCF",
                                      "DEP", "PRED", "GRAZ", "SCAV", "PARA", "UNK"))
  )

# ============ 6. CREATE MULTI-PANEL FIGURE ============

cat("\nGenerating figure panels...\n")

# PANEL A: Stacked bar chart
p1 <- ggplot(composition_long, aes(x = Site, y = Proportion, fill = Functional_Group)) +
  geom_bar(stat = "identity", color = "white", size = 0.3) +
  scale_fill_manual(values = functional_colors, 
                   name = "Functional Group",
                   labels = c("PHOT" = "Photoautotroph",
                            "FDM" = "Fine Dead-End Microfilterer",
                            "ADS" = "Active Dead-End Sieve",
                            "WCP" = "Weak Ciliary Pumper",
                            "TPC" = "Truly Passive Collector",
                            "RNM" = "Renewable Mucus Net",
                            "RCF" = "Ram/Crossflow",
                            "DEP" = "Deposit Feeder",
                            "PRED" = "Predator",
                            "GRAZ" = "Grazer",
                            "SCAV" = "Scavenger",
                            "PARA" = "Parasite",
                            "UNK" = "Unclassified")) +
  scale_x_discrete(labels = function(x) gsub("_", "\n", x)) +
  labs(x = NULL, y = "Relative Abundance (%)", 
       title = "A) Functional Composition Across Turbidity Gradient") +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm"),
    plot.title = element_text(face = "bold", size = 11)
  )

# PANEL B: Functional richness vs turbidity
functional_richness <- composition_long %>%
  filter(Proportion > 1) %>%
  mutate(Site = as.character(Site)) %>%  # Convert factor to character
  group_by(Site) %>%
  dplyr::summarize(Richness = n()) %>%
  left_join(
    functional_composition %>% 
      mutate(Site = as.character(Site)) %>%  # Convert to character
      dplyr::select(Site, Turbidity) %>%
      dplyr::rename(Turbidity_value = Turbidity),
    by = "Site"
  ) %>%
  dplyr::mutate(Site = factor(Site, levels = turb_ordered_sites))

p2 <- ggplot(functional_richness, aes(x = Turbidity_value, y = Richness)) +
  geom_smooth(method = "lm", se = TRUE, color = "black", size = 0.8, alpha = 0.2) +
  geom_point(aes(fill = Site), size = 4, shape = 21, color = "black") +
  scale_fill_manual(values = c12_sites) +
  scale_x_reverse(limits = c(12, 0)) +  # Reversed limits for reversed scale
  labs(x = "Turbidity (NTU)", y = "Functional Group Richness",
       title = "B) Functional Diversity Decline") +
  theme_classic(base_size = 11) +
  theme(legend.position = "none", plot.title = element_text(face = "bold", size = 11))

# PANEL C - with character conversion
key_groups <- composition_long %>%
  filter(Functional_Group %in% c("FDM", "ADS", "PHOT")) %>%
  mutate(Site = as.character(Site)) %>%
  left_join(
    functional_composition %>% 
      mutate(Site = as.character(Site)) %>%
      dplyr::select(Site, Turbidity) %>%
      dplyr::rename(Turbidity_value = Turbidity),
    by = "Site"
  )

p3 <- ggplot(key_groups, aes(x = Turbidity_value, y = Proportion, color = Functional_Group)) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2, size = 1) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = functional_colors, 
                     name = "Functional Group",
                     labels = c("FDM" = "FDM (Sponges)",
                              "ADS" = "ADS (Bivalves/Barnacles)",
                              "PHOT" = "Photoautotrophs")) +
  scale_x_reverse(limits = c(12, 0)) +  # Corrected
  labs(x = "Turbidity (NTU)", y = "Relative Abundance (%)",
       title = "C) Key Groups: Predicted vs Observed") +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 11))

# PANEL D - with character conversion
functional_shannon <- composition_long %>%
  mutate(Site = as.character(Site)) %>%
  group_by(Site) %>%
  mutate(p = Proportion / 100) %>%
  dplyr::summarize(Shannon = -sum(p * log(p + 1e-10), na.rm = TRUE)) %>%
  left_join(
    functional_composition %>% 
      mutate(Site = as.character(Site)) %>%
      dplyr::select(Site, Turbidity) %>%
      dplyr::rename(Turbidity_value = Turbidity),
    by = "Site"
  ) %>%
  mutate(Site = factor(Site, levels = turb_ordered_sites))

p4 <- ggplot(functional_shannon, aes(x = Turbidity_value, y = Shannon)) +
  geom_smooth(method = "lm", se = TRUE, color = "black", size = 0.8, alpha = 0.2) +
  geom_point(aes(fill = Site), size = 4, shape = 21, color = "black") +
  scale_fill_manual(values = c12_sites) +
  scale_x_reverse(limits = c(12, 0)) +  # Corrected
  labs(x = "Turbidity (NTU)", y = "Shannon Diversity (H')",
       title = "D) Functional Evenness Loss") +
  theme_classic(base_size = 11) +
  theme(legend.position = "none", plot.title = element_text(face = "bold", size = 11))


# Now combine and print
combined_figure <- (p1 | (p2 / p4)) / p3 +
  plot_layout(heights = c(1.2, 1))

print(combined_figure)

# Save figure
ggsave(file.path(output_dir, "functional_composition_updated.pdf"), 
       combined_figure, width = 14, height = 10, units = "in")

# ============ 7. SUMMARY STATISTICS ============

# === SUMMARY STATISTICS ===

# Correlation tests - use Turbidity_value
cor_richness <- cor.test(functional_richness$Turbidity_value, functional_richness$Richness)
cat("Functional richness vs turbidity:\n")
cat("  r =", round(cor_richness$estimate, 3), "\n")
cat("  p =", format.pval(cor_richness$p.value, digits = 3), "\n\n")

cor_shannon <- cor.test(functional_shannon$Turbidity_value, functional_shannon$Shannon)
cat("Functional diversity vs turbidity:\n")
cat("  r =", round(cor_shannon$estimate, 3), "\n")
cat("  p =", format.pval(cor_shannon$p.value, digits = 3), "\n\n")

# Mean proportions by turbidity category - also use Turbidity_value
functional_by_turbidity <- functional_composition %>%
  dplyr::mutate(
    Site = as.character(Site),
    Turbidity_value = Turbidity  # Create this column if using original Turbidity
  ) %>%
  dplyr::mutate(Turbidity_Category = ifelse(Turbidity_value > 6, "High", "Low")) %>%
  group_by(Turbidity_Category) %>%
  dplyr::summarize(
    across(any_of(c("PHOT", "FDM", "ADS", "WCP", "DEP")),
           ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

cat("Mean functional composition by turbidity category:\n")
print(functional_by_turbidity)

# Export composition data
write.csv(functional_composition, 
          file.path(output_dir, "functional_composition_by_site.csv"),
          row.names = FALSE)
```


```{r}
# Check phylum composition of PHOT
phot_phyla <- functional_dataset %>%
  dplyr::filter(is_photoautotroph == 1, as.logical(Is_Significant), Fraction == "All_Samples") %>%
  dplyr::count(final_phylum) %>%
  dplyr::arrange(desc(n))

print(phot_phyla)
```

