---
title: "seascape_18S"
author: "Nicole Pittoors"
date: "2024-09-10"
output: html_document
---
#Set working directory

setwd("Path/to/env_data")

Load libraries
```{r}
library(ncdf4)
library(lubridate)
library(RColorBrewer)
library(lattice)
library(dplyr)
library(oce)
library(raster)
library(seacarb)
library(rnaturalearth) # library to add polygons of the countries to a map
library(rnaturalearthdata) # library to add polygons of the countries to a map
```

Import geographic coordinates files
```{r}
#by site
# we change the order of the columns, as longitude generally is the first column and latitude the second one.
coord_sites <-data.frame(row.names=unique(metadata_ARMS$sitelocality))
b<-numeric(length(unique(metadata_ARMS$sitelocality)))
c<-numeric(length(unique(metadata_ARMS$sitelocality)))
for(i in 1:length(rownames(coord_sites))){
 b[i]<-mean(metadata_ARMS$latitude[metadata_ARMS$sitelocality==rownames(coord_sites)[i]])
  c[i]<-mean(metadata_ARMS$longitude[metadata_ARMS$sitelocality==rownames(coord_sites)[i]])
}
coord_sites$longitude<-c
coord_sites$latitude<-b
#by individual
coord_ind <- metadata_ARMS[,c("longitude","latitude")]
# we can visualize the content of the table:
head(coord_sites)
head(coord_ind)
```

========================
Process Oceancolour data
========================
Open optics NetDCF file downloaded from Oceancolour
Variables: 
1) "BBP" = volume_backwards_scattering_coefficient_of_radiative_flux_in_sea_water_due_to_particles
  Particulate back-scattering coefficient at 443 nm - Mean of the binned pixels float BBP_uncertainty[longitude,latitude,time]
2) "BBP_uncertainty" = volume_absorption_coefficient_of_radiative_flux_in_sea_water_due_to_dissolved_organic_matter_and_non_algal_particles
  Particulate back-scattering coefficient at 443 nm - Uncertainty estimation float CDM[longitude,latitude,time]
3) "CDM" = Coloured dissolved and detrital organic materials - Mean of the binned pixels float CDM_uncertainty[longitude,latitude,time]
4) "CDM_uncertainty" = Coloured dissolved and detrital organic materials - Uncertainty estimation float flags[longitude,latitude,time]
5) status flags
```{r}
nc_file_optics <- nc_open("cmems_obs-oc_glo_bgc-optics_my_l4-multi-4km_P1M_1725914557874.nc")
print(nc_file)
```

Open plankton NetDC file downloaded from Oceancolour
Variables:
1) mass_concentration_of_chlorophyll_a_in_sea_water
  Chlorophyll-a concentration - Mean of the binned pixels
2) Chlorophyll-a concentration - Uncertainty estimation
```{r}
nc_file_plankton <- nc_open("cmems_obs-oc_glo_bgc-plankton_my_l4-multi-4km_P1M_1725914300477.nc")
print(nc_file_plankton)
```

Open primary productivity (PP) NetDC file downloaded from Oceancolour
Variables:
1) primary_productivity_of_biomass_expressed_as_carbon
  Primary Productivity - Mean of the binned pixels
2) Primary Productivity - Uncertainty estimation
3) status_flag
```{r}
nc_file_PP <- nc_open("cmems_obs-oc_glo_bgc-pp_my_l4-multi-4km_P1M_1725914400069.nc")
print(nc_file_PP)
```

Open transp NetDC file downloaded from Oceancolour
Open primary productivity (PP) NetDC file downloaded from Oceancolour
Variables:
1) KD490 = Volume attenuation coefficient of downwelling radiative flux in sea water
2) KD490_uncertainty = Diffuse attenuation coefficient - Uncertainty estimation
3) SPM (g/m^3) = Mass concentration of suspended matter in sea water
4) SPM_uncertainty (%)= Inorganic suspended particulate matter in sea water - Uncertainty estimation
5) ZSD (m) = Secchi depth of sea water
6) ZSD_uncertainty (%) = Secchi disk depth - Uncertainty estimation
7) flags = status_flag
```{r}
nc_file_transp <- nc_open("cmems_obs-oc_glo_bgc-transp_my_l4-multi-4km_P1M_1726252158404.nc")
print(nc_file_transp)
```


Surface primary productivity (_PP)
1) Convert time variable, current time is seconds since a 1900s date
```{r}
# Extract time variable and units from the NetCDF file
time <- nc_file_PP$dim[[3]]$vals
t_units_PP <- ncatt_get(nc_file_PP, "time", "units")

# Extract the reference date from the units string (should be something like "seconds since 1970-01-01 00:00:00")
# The third element contains the reference date (split by space)
t_ustr <- strsplit(t_units_PP$value, " ")[[1]]
reference_date <- ymd_hms(paste(t_ustr[3], t_ustr[4]))  # This will be "1970-01-01 00:00:00" by default

# Convert the time from seconds since the reference date
date <- reference_date + seconds(time)

# Print the results
date
```

2) Convert the .nc file to .tiff while adding the 'date' attribute that was just created
```{r}
#Extract the variables
PP_data <- ncvar_get(nc_file_PP, "PP")
lat <- ncvar_get(nc_file_PP, "latitude")
lon <- ncvar_get(nc_file_PP, "longitude")
time <- ncvar_get(nc_file_PP, "time")

#Convert time to date
time_units <- ncatt_get(nc_file_PP, "time", "units")
origin <- sub("seconds since ", "", time_units$value)
dates <- as.POSIXct(time, origin=origin, tz="UTC")

#Create a RasterBrick Object- combine the PP data and the coordinates into a raster stack
PP_raster <- brick(PP_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +datum=WGS84"))

# Write the Raster to a GeoTIFF
writeRaster(PP_raster, "PP_data.tif", format="GTiff",overwrite = TRUE )
```

Calculate the mean and standard deviation of PP at sampling sites by extracting the PP values from the raster for each site in coord_sites dataframe following these three steps:
	1.	Extract PP Values for Sampling Sites: Extract PP data from the raster at the coordinates from coord_sites.
	2.	Calculate Mean and Standard Deviation: Compute the statistics for each site over time.
	3.	Store Results: Store the results for mean and standard deviation in your metadata_ARMS object or a new object.
```{r}
# Load necessary libraries
#library(raster)

# Loop through each site to extract the data for mean and standard deviation calculation
for (site in rownames(coord_sites)) {
  
  # Get the corresponding latitude and longitude for the site
  site_lat <- coord_sites[site, "latitude"]
  site_lon <- coord_sites[site, "longitude"]
  
  # Find the nearest raster cell to the site's coordinates
  cell_index <- cellFromXY(PP_raster, c(site_lon, site_lat))
  
  # Extract the CHL time series data for this specific cell (site)
  PP_site_data <- extract(PP_raster, cell_index, df = TRUE)
  
  # Calculate mean and standard deviation across the time dimension
  PP_site_mean <- apply(PP_site_data[, -1], 1, mean, na.rm = TRUE)
  PP_site_sd <- apply(PP_site_data[, -1], 1, sd, na.rm = TRUE)
  
  # Add the extracted data back to your metadata_ARMS or store in a new dataframe
  metadata_ARMS[metadata_ARMS$sitelocality == site, "PP_mean"] <- PP_site_mean
  metadata_ARMS[metadata_ARMS$sitelocality == site, "PP_sd"] <- PP_site_sd
}

# Print the updated metadata to check if values are stored correctly
head(metadata_ARMS)

# Optionally, save your updated metadata if necessary
#write.csv(metadata_ARMS, "updated_metadata.csv", row.names = FALSE)

# Confirm the process
#print("PP mean and standard deviation added to metadata_ARMS.")
```

Chlorophyll A Concentration PLANKTON CHL
1) Convert time variable, current time is seconds since a 1900s date
```{r}
# Extract time variable and units from the NetCDF file
time <- nc_file_plankton$dim[[3]]$vals
t_units_plankton <- ncatt_get(nc_file_plankton, "time", "units")

# Extract the reference date from the units string (should be something like "seconds since 1970-01-01 00:00:00")
# The third element contains the reference date (split by space)
t_ustr <- strsplit(t_units_plankton$value, " ")[[1]]
reference_date <- ymd_hms(paste(t_ustr[3], t_ustr[4]))  # This will be "1970-01-01 00:00:00" by default

# Convert the time from seconds since the reference date
date <- reference_date + seconds(time)

# Print the results
date
```

2) Convert the .nc file to .tiff while adding the 'date' attribute that was just created
```{r}
#Extract the variables
CHL_data <- ncvar_get(nc_file_plankton, "CHL")
lat <- ncvar_get(nc_file_plankton, "latitude")
lon <- ncvar_get(nc_file_plankton, "longitude")
time <- ncvar_get(nc_file_plankton, "time")

#Convert time to date
time_units <- ncatt_get(nc_file_plankton, "time", "units")
origin <- sub("seconds since ", "", time_units$value)
dates <- as.POSIXct(time, origin=origin, tz="UTC")

#Create a RasterBrick Object- combine the CHL data and the coordinates into a raster stack
CHL_raster <- brick(CHL_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +datum=WGS84"))

# Write the Raster to a GeoTIFF
writeRaster(CHL_raster, "CHL_data.tif", format="GTiff",overwrite = TRUE )
```

Calculate the mean and standard deviation of CHL at sampling sites by extracting the CHL values from the raster for each site in coord_sites dataframe following these three steps:
	1.	Extract CHL Values for Sampling Sites: Extract CHL data from the raster at the coordinates from coord_sites.
	2.	Calculate Mean and Standard Deviation: Compute the statistics for each site over time.
	3.	Store Results: Store the results for mean and standard deviation in your metadata_ARMS object or a new object.
```{r}
# Load necessary libraries
#library(raster)

# Loop through each site to extract the data for mean and standard deviation calculation
for (site in rownames(coord_sites)) {
  
  # Get the corresponding latitude and longitude for the site
  site_lat <- coord_sites[site, "latitude"]
  site_lon <- coord_sites[site, "longitude"]
  
  # Find the nearest raster cell to the site's coordinates
  cell_index <- cellFromXY(CHL_raster, c(site_lon, site_lat))
  
  # Extract the CHL time series data for this specific cell (site)
  CHL_site_data <- extract(CHL_raster, cell_index, df = TRUE)
  
  # Calculate mean and standard deviation across the time dimension
  CHL_site_mean <- apply(CHL_site_data[, -1], 1, mean, na.rm = TRUE)
  CHL_site_sd <- apply(CHL_site_data[, -1], 1, sd, na.rm = TRUE)
  
  # Add the extracted data back to your metadata_ARMS or store in a new dataframe
  metadata_ARMS[metadata_ARMS$sitelocality == site, "CHL_mean"] <- CHL_site_mean
  metadata_ARMS[metadata_ARMS$sitelocality == site, "CHL_sd"] <- CHL_site_sd
}

# Print the updated metadata to check if values are stored correctly
head(metadata_ARMS)

# Optionally, save your updated metadata if necessary
write.csv(metadata_ARMS, "./env_dataPP_metadata_ARMS.csv", row.names = FALSE)

# Confirm the process
#print("CHL mean and standard deviation added to metadata_ARMS.")
```

Inorganic suspended particulates TRANSP SPM
1) Convert time variable, current time is seconds since a 1900s date
```{r}
# Extract time variable and units from the NetCDF file
time <- nc_file_transp$dim[[3]]$vals
t_units_transp <- ncatt_get(nc_file_transp, "time", "units")

# Extract the reference date from the units string (should be something like "seconds since 1970-01-01 00:00:00")
# The third element contains the reference date (split by space)
t_ustr <- strsplit(t_units_transp$value, " ")[[1]]
reference_date <- ymd_hms(paste(t_ustr[3], t_ustr[4]))  # This will be "1970-01-01 00:00:00" by default

# Convert the time from seconds since the reference date
date <- reference_date + seconds(time)

# Print the results
date
```

2) Convert the .nc file to .tiff while adding the 'date' attribute that was just created
```{r}
#Extract the variables
SPM_data <- ncvar_get(nc_file_transp, "SPM")
lat <- ncvar_get(nc_file_transp, "latitude")
lon <- ncvar_get(nc_file_transp, "longitude")
time <- ncvar_get(nc_file_transp, "time")

#Convert time to date
time_units <- ncatt_get(nc_file_transp, "time", "units")
origin <- sub("seconds since ", "", time_units$value)
dates <- as.POSIXct(time, origin=origin, tz="UTC")

#Create a RasterBrick Object- combine the SPM data and the coordinates into a raster stack
SPM_raster <- brick(SPM_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +datum=WGS84"))

# Write the Raster to a GeoTIFF
writeRaster(SPM_raster, "SPM_data.tif", format="GTiff",overwrite = TRUE )
```

3) Calculate the mean and standard deviation of SPM at sampling sites by extracting the SPM values from the raster for each site in coord_sites
```{r}
# Loop through each site to extract the data for mean and standard deviation calculation
for (site in rownames(coord_sites)) {
  
  # Get the corresponding latitude and longitude for the site
  site_lat <- coord_sites[site, "latitude"]
  site_lon <- coord_sites[site, "longitude"]
  
  # Find the nearest raster cell to the site's coordinates
  cell_index <- cellFromXY(SPM_raster, c(site_lon, site_lat))
  
  # Extract the CHL time series data for this specific cell (site)
  SPM_site_data <- extract(SPM_raster, cell_index, df = TRUE)
  
  # Calculate mean and standard deviation across the time dimension
  SPM_site_mean <- apply(SPM_site_data[, -1], 1, mean, na.rm = TRUE)
  SPM_site_sd <- apply(SPM_site_data[, -1], 1, sd, na.rm = TRUE)
  
  # Add the extracted data back to your metadata_ARMS or store in a new dataframe
  metadata_ARMS[metadata_ARMS$sitelocality == site, "SPM_OM"] <- SPM_site_mean
  metadata_ARMS[metadata_ARMS$sitelocality == site, "SPM_OSD"] <- SPM_site_sd
}

# Print the updated metadata to check if values are stored correctly
head(metadata_ARMS)
```


Particulate back scattering coefficient OPTICS
1) Convert time variable, current time is seconds since a 1900s date
```{r}
# Extract time variable and units from the NetCDF file
time <- nc_file_optics$dim[[3]]$vals
t_units_optics <- ncatt_get(nc_file_optics, "time", "units")

# Extract the reference date from the units string (should be something like "seconds since 1970-01-01 00:00:00")
# The third element contains the reference date (split by space)
t_ustr <- strsplit(t_units_optics$value, " ")[[1]]
reference_date <- ymd_hms(paste(t_ustr[3], t_ustr[4]))  # This will be "1970-01-01 00:00:00" by default

# Convert the time from seconds since the reference date
date <- reference_date + seconds(time)

# Print the results
date
```

2) Convert the .nc file to .tiff while adding the 'date' attribute that was just created
```{r}
#Extract the variables
bbp_data <- ncvar_get(nc_file_optics, "BBP")
lat <- ncvar_get(nc_file_optics, "latitude")
lon <- ncvar_get(nc_file_optics, "longitude")
time <- ncvar_get(nc_file_optics, "time")

#Convert time to date
time_units <- ncatt_get(nc_file_optics, "time", "units")
origin <- sub("seconds since ", "", time_units$value)
dates <- as.POSIXct(time, origin=origin, tz="UTC")

#Create a RasterBrick Object- combine the BBP data and the coordinates into a raster stack
BBP_raster <- brick(bbp_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +datum=WGS84"))

# Write the Raster to a GeoTIFF
writeRaster(BBP_raster, "BBP_data.tif", format="GTiff",overwrite = TRUE )
```

3) Calculate the mean and standard deviation of BBP at sampling sites by extracting the BBP values from the raster for each site in coord_sites
```{r}
# Loop through each site to extract the data for mean and standard deviation calculation
for (site in rownames(coord_sites)) {
  
  # Get the corresponding latitude and longitude for the site
  site_lat <- coord_sites[site, "latitude"]
  site_lon <- coord_sites[site, "longitude"]
  
  # Find the nearest raster cell to the site's coordinates
  cell_index <- cellFromXY(BBP_raster, c(site_lon, site_lat))
  
  # Extract the CHL time series data for this specific cell (site)
  BBP_site_data <- extract(BBP_raster, cell_index, df = TRUE)
  
  # Calculate mean and standard deviation across the time dimension
  BBP_site_mean <- apply(BBP_site_data[, -1], 1, mean, na.rm = TRUE)
  BBP_site_sd <- apply(BBP_site_data[, -1], 1, sd, na.rm = TRUE)
  
  # Add the extracted data back to your metadata_ARMS or store in a new dataframe
  metadata_ARMS[metadata_ARMS$sitelocality == site, "BBP_OM"] <- BBP_site_mean
  metadata_ARMS[metadata_ARMS$sitelocality == site, "BBP_OSD"] <- BBP_site_sd
}

# Print the updated metadata to check if values are stored correctly
head(metadata_ARMS)
```

CDM OPTICS
1) Convert time variable, current time is seconds since a 1900s date
```{r}
# Extract time variable and units from the NetCDF file
time <- nc_file_optics$dim[[3]]$vals
t_units_optics <- ncatt_get(nc_file_optics, "time", "units")

# Extract the reference date from the units string (should be something like "seconds since 1970-01-01 00:00:00")
# The third element contains the reference date (split by space)
t_ustr <- strsplit(t_units_optics$value, " ")[[1]]
reference_date <- ymd_hms(paste(t_ustr[3], t_ustr[4]))  # This will be "1970-01-01 00:00:00" by default

# Convert the time from seconds since the reference date
date <- reference_date + seconds(time)

# Print the results
date
```

2) Convert the .nc file to .tiff while adding the 'date' attribute that was just created
```{r}
#Extract the variables
CDM_data <- ncvar_get(nc_file_optics, "CDM")
lat <- ncvar_get(nc_file_optics, "latitude")
lon <- ncvar_get(nc_file_optics, "longitude")
time <- ncvar_get(nc_file_optics, "time")

#Convert time to date
time_units <- ncatt_get(nc_file_optics, "time", "units")
origin <- sub("seconds since ", "", time_units$value)
dates <- as.POSIXct(time, origin=origin, tz="UTC")

#Create a RasterBrick Object- combine the CDM data and the coordinates into a raster stack
CDM_raster <- brick(CDM_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +datum=WGS84"))

# Write the Raster to a GeoTIFF
writeRaster(CDM_raster, "CDM_data.tif", format="GTiff",overwrite = TRUE )
```

3) Calculate the mean and standard deviation of CDM at sampling sites by extracting the CDM values from the raster for each site in coord_sites
```{r}
# Loop through each site to extract the data for mean and standard deviation calculation
for (site in rownames(coord_sites)) {
  
  # Get the corresponding latitude and longitude for the site
  site_lat <- coord_sites[site, "latitude"]
  site_lon <- coord_sites[site, "longitude"]
  
  # Find the nearest raster cell to the site's coordinates
  cell_index <- cellFromXY(CDM_raster, c(site_lon, site_lat))
  
  # Extract the CHL time series data for this specific cell (site)
  CDM_site_data <- extract(CDM_raster, cell_index, df = TRUE)
  
  # Calculate mean and standard deviation across the time dimension
  CDM_site_mean <- apply(CDM_site_data[, -1], 1, mean, na.rm = TRUE)
  CDM_site_sd <- apply(CDM_site_data[, -1], 1, sd, na.rm = TRUE)
  
  # Add the extracted data back to your metadata_ARMS or store in a new dataframe
  metadata_ARMS[metadata_ARMS$sitelocality == site, "CDM_OM"] <- CDM_site_mean
  metadata_ARMS[metadata_ARMS$sitelocality == site, "CDM_OSD"] <- CDM_site_sd
}

# Print the updated metadata to check if values are stored correctly
head(metadata_ARMS)
```

3) Close the Net CDF file
```{r}
nc_close(nc_file)
```


============================
Load Environmental Matrix
============================

```{r}
EnvMatrix <- read.csv("./EnMatrix_site.csv",row.names=1, sep=",")

# Convert temp, salinity, and depth columns to numeric
EnvMatrix$temp <- as.numeric(EnvMatrix$temp)
EnvMatrix$salinity <- as.numeric(EnvMatrix$salinity)
EnvMatrix$depth <- as.numeric(EnvMatrix$depth)

# Check if the variables were successfully converted
str(EnvMatrix)
```


==============================================
Summarize patterns of environmental variation
==============================================

1) Perform a Principal Component analysis of the environmental matrix
- A principal component analysis (PCA) is a statistical method to summarize the complexity of a numerical matrix.
- Applied to an environmental matrix, for example, the PCA informs on 1) which sampling sites show similar environmental conditions  and 2) which environmental conditions are correlated (i.e. show similar spatial patterns) 

Plot and run correlation test on environmental variables
```{r}
library(psych)
library(Hmisc)
library(corrplot)
pairs.panels(EnvMatrix, scale = TRUE)
site.rcorr = rcorr(as.matrix(EnvMatrix))
site.coeff = site.rcorr$r
site.p = site.rcorr$P
corrplot(site.coeff,p.mat=site.p)
```

```{r}
# Load the correlation coefficients from CSV
site_coeff_from_csv <- read.csv("./site_coefficients.csv", row.names = 1)

# Convert to matrix if it's loaded as a data frame
site_coeff_matrix <- as.matrix(site_coeff_from_csv)

# Create a basic correlation plot
corplot<- corrplot(site_coeff_matrix, method = "color", 
         tl.col = "black", # Text label color
         addCoef.col = "black") # Add coefficient values

pdf("correlation_plot_high_res.pdf", width = 16, height = 12)
corrplot(site_coeff_matrix, method = "color", 
         tl.col = "black",
         addCoef.col = "black",
         mar = c(1, 1, 1, 1))  # Adjust margins if needed
dev.off()
```



```{r}
# Load necessary libraries
#library(corrplot)

# Define function to calculate p-values for correlations
cor.mtest <- function(mat, conf.level = 0.95) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p.mat <- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], conf.level = conf.level)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

# Load your environmental matrix (replace with actual data loading code)
# EnvMatrix_site <- read.csv("your_file.csv")  # Example: Load from CSV

# Compute correlation matrix
site.coeff <- cor(EnvMatrix, use = "pairwise.complete.obs", method = "pearson")

# Compute p-values
site.p <- cor.mtest(EnvMatrix)

# Plot the correlation matrix
corrplot(site.coeff, p.mat = site.p, sig.level = 0.05, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45, number.cex = 0.7)
```

```{r}
# Install and load ggcorrplot
# install.packages("ggcorrplot")
library(ggcorrplot)

# Plot correlation matrix with ggcorrplot
env_cor_plot<- ggcorrplot(site.coeff, method = "circle", type = "upper", p.mat = site.p, 
           sig.level = 0.05, lab = TRUE)

# Write correlation matrix (site.coeff) to .csv
write.csv(site.coeff, file = "./site_coefficients.csv", row.names = TRUE)
```

Subset data to non-highly correlated variables
```{r}
EnvMatrix_reduced = subset (EnvMatrix, select = c(salinity,depth,trubidity_std_rank,longitude,PP_mean,temp_HOBO_OM))
pairs.panels(EnvMatrix_reduced, scale = TRUE)
```
EnvMatrix_reduced
salinity depth trubidity_std_rank longitude    SPM_OM   PP_mean temp_HOBO_OM
Alderdice_coral      36.41000    82          11.285714 -92.00577 0.3505688  607.4360     20.74898
Alderdice_background 36.44000    82          11.714286 -92.00566 0.3505688  607.4360     20.66798
Alderdice_shallow    36.31667    66           3.285714 -91.99257 0.3519777  606.3403     22.07339
Bright_coral         36.29200    83           7.428571 -93.27030 0.3537832  417.8301     20.60550
Bright_background    36.29333    83           6.857143 -93.27022 0.3537832  417.8301     20.73661
Bright_shallow       36.31182    65           4.142857 -93.27173 0.3528762  417.8301     21.64840
Diaphus_background   35.49667    82           5.285714 -90.70140 0.3543301 1130.9303     20.97211
Diaphus_coral        35.34000    81           6.428571 -90.70141 0.3543301 1130.9303     20.99291
EFGB_shallow         36.28000    55           2.428571 -93.60958 0.3484201  395.8951     22.55360
EFGB_deep            36.13667    81           5.857143 -93.62013 0.3526149  395.8951     20.76290
McGrail_coral        36.43333    85           2.000000 -92.58129 0.3494234  502.7808     20.62179
Stetson_coral        36.22833    54          11.285714 -94.30297 0.3492222  378.0407     22.14335


Save .Rdata
```{r}
save(EnvMatrix_reduced,file='./EnvMatrix_reduced.Rdata')
```

Apply scale transformation to every column of the matrix. 
- This is necessary since the downstream analysis is sensitive to the unit scales of the different environmental variables
- By using the scale function, all the environmental variables are standardized to an identical scale
```{r}
sEnvMatrix = scale(EnvMatrix_reduced, center = T, scale = T)
```

Take a look at what the dataset looked like before the standardization of the unites..
```{r}
boxplot(EnvMatrix_reduced)
```

.. .and after scale transformation
```{r}
boxplot(sEnvMatrix)
```

Compute a heatmap of the environmental matrix to understand why we need a PCA
Columns correspond to different environmental variables, the rows to the different sampling locations
The heatmap highlights similarities between variables and between sampling sites.
```{r}
heatmap(sEnvMatrix)
```

# Creating figures for publication
```{r}
# Load the CSV file
env_data <- read.csv("./env_data/EnMatrix_site.csv", row.names = 1)

env_matrix <- as.matrix(env_data)

# Create the yellow-orange-red color palette
heat_colors <- colorRampPalette(c("#FFFFCC", "#FED976", "#FD8D3C", "#E31A1C", "#800026"))(100)

# Create full heatmap with all variables
pdf("all_variables_heatmap.pdf", width = 12, height = 10)
heatmap(env_matrix, 
        scale = "column",  
        col = heat_colors,  
        margins = c(10, 8),
        main = "Heatmap of All Environmental Variables")
dev.off()

# For the selected variables, use the correct column names from your data
selected_vars <- env_matrix[, c("trubidity_std_rank", "depth", "PP_mean")]

pdf("selected_variables_heatmap.pdf", width = 8, height = 10)
heatmap(selected_vars, 
        scale = "column",
        col = heat_colors,
        margins = c(5, 8),
        main = "Heatmap of Selected Variables")
dev.off()
```


The goal of the PCA is to create proxy variables that represent the combination of correlated variables. 
These proxy variables are called principal components. 
By calculating principal components, we will be able to describe the environmental variation of the matrix using less variables. 
We can use the prcomp() function to perform the principal component analysis.
```{r}
ePCA = prcomp(sEnvMatrix)

# The principal components can be found in the $x matrix
ePCA$x
```

Check the correlation of principal components are correlated. They should be non-correlated and therefore describe non-overlapping patterns of overlapping data.
```{r}
cor(ePCA$x)
```

The principal components are sorted based on the % of variance that they can explain. 
Plot the first two PCs
This plot informs us that according to the main axis of environmental variation (PC1), sites
```{r}
plot(ePCA$x[,1:2], cex=0.5)
text(ePCA$x[,1:2], labels = rownames(EnvMatrix_reduced)) # add label of sampling site identifier 
```

What does this main axis of variation mean? We can answer this by looking at another matrix stored in the ePCA object: the rotation matrix
This matrix informs us on which environmental variables participate to the composition of each principal components. The contribution of each variable is called the "loading"
```{r}
ePCA$rotation
```

We can visualize this information for PC1:
```{r}
barplot(ePCA$rotation[,'PC1'], las =2, main='PC1-loadings')
```

for PC2:
```{r}
barplot(ePCA$rotation[,'PC2'], las =2, main='PC2-loadings')
```

PC1 is just one of many principal components, each explaining a different contrast of environmental conditions. 
We can quantify the % of total variation explained by PC1 by looking at the summary of the PCA.
The second row describes the proportion of variance explained by each principal component.
```{r}
S=summary(ePCA)
S
```

We can therefore plot the % explained by each component as
```{r}
plot(S$importance[2,], ylab='% of explained variance', xlab='#PC')
```

WE can also visualize this as a cumulative plot (he third row of the summary)
```{r}
plot(S$importance[3,], ylab='% of explained variance', xlab='#PC', ylim=c(0,1))
abline(h=0.9)
```

===========================================
Hierarchical clustering
===========================================
Load CSV with environmental measurements (reduced) for each ARMS, this is necessary for comparing geographic and environmental distance to bray-curtis dissimilarity. 08/22/25- Corrected SH ID to corresponding env values
```{r}
EnvMatrix_ind <- read.csv("./EnvMatrix_ind_corrected.csv",row.names=1, sep=",")
# Convert to dataframe
EnvMatrix_ind <- as.data.frame(EnvMatrix_ind)
# Scale transform around zero
sEnvMatrix_ind = scale(EnvMatrix_ind, center = T, scale = T)
boxplot(sEnvMatrix_ind)
```



##Calculating Environmental Distances

1) Define the environmental variables we want to use to define the (dis-)similarity between sites and calculate the euclidean environmental distance based on these variables:
"salinity", "depth", "trubidity_std_rank", "longitude", "PP_mean", "temp_HOBO_OM"   
```{r}
depth_dist <- dist(EnvMatrix_ind$depth)
turb_dist <- dist(EnvMatrix_ind$turbidity_std_rank)
PP_dist <- dist(EnvMatrix_ind$PP_mean)
#temp_dist <- dist(EnvMatrix_ind$temp_mean)
```

2) Calculate the environmental distances while considering all the environmental variables in the dataset
You MUST use the scaled variables
```{r}
env_dist <- dist(sEnvMatrix_ind)
env_dist
# Convert the distance matrix to a data frame
env_dist_df <- as.data.frame(as.matrix(env_dist))
# Write the data frame to a CSV file to view full matrix output
write.csv(env_dist_df, "./env_dist_matrix.csv", row.names = TRUE)
```

3) Calculate environmnetal distances based on some of the principal components SKIPPING
```{r}
dist(ePCA$x[1,2])
```


#Calculate environmental distances for sessile fraction samples
```{r}
# Subset the EnMatrix_ind dataframe
EnvMatrix_sessile <- EnvMatrix_ind[rownames(EnvMatrix_ind) %in% sessile_rows, ]

# View the subset
head(EnvMatrix_sessile)

sEnvMatrix_sessile = scale(EnvMatrix_sessile, center = T, scale = T) # Scale transform
sEnvMatrix_sessile <- as.data.frame(sEnvMatrix_sessile) # Convert to dataframe
env_dist_sess <- dist(sEnvMatrix_sessile) # Calculate euclidean distances
boxplot(sEnvMatrix_sessile) # Plot scale transformed values

```

Calculate motile env distances
```{r}
sEnvMatrix_motile = scale(EnvMatrix_motile, center = T, scale = T) # Scale transform
sEnvMatrix_motile <- as.data.frame(sEnvMatrix_motile) # Convert to dataframe
env_dist_mot <- dist(sEnvMatrix_motile) # Calculate euclidean distances
boxplot(sEnvMatrix_motile) # Plot scale transformed values
```

# Reduce dimensions of environmental matrices for motile and sessile fractions
Need to reduce dimensions of environmental matrix (114x114 to 78x78, subset using SH number)
1) Call rownames (SH #s) that are 100 or 500 in fraction variable in metadata_ARMS
```{r}
# Motile fraction -Print the row names where 'fraction' equals 100 or 500
rownames(metadata_ARMS)[which(metadata_ARMS$fraction %in% c(100, 500))]
rownames(metadata_ARMS)[which(metadata_ARMS$fraction %in% c("sessile"))]

# Define the vector of motile rows
motile_rows <- c("SH21748", "SH21749", "SH21750", "SH21751", "SH21752", "SH21753", 
                   "SH21754", "SH21755", "SH21756", "SH21757", "SH21758", "SH21759", 
                   "SH21760", "SH21761", "SH21762", "SH21763", "SH21764", "SH21765", 
                   "SH21772", "SH21773", "SH21774", "SH21776", "SH21775", "SH21777", 
                   "SH21778", "SH21779", "SH21780", "SH21781", "SH21782", "SH21783", 
                   "SH21784", "SH21785", "SH21786", "SH21787", "SH21789", "SH21790", 
                   "SH21788", "SH21791", "SH21792", "SH21736", "SH21737", "SH21738", 
                   "SH21739", "SH21740", "SH21741", "SH21742", "SH21743", "SH21744", 
                   "SH21745", "SH21746", "SH21747", "SH21802", "SH21803", "SH21804", 
                   "SH21805", "SH21806", "SH21807", "SH21808", "SH21809", "SH21810", 
                   "SH21811", "SH21812", "SH21813", "SH21766", "SH21767", "SH21768", 
                   "SH21769", "SH21770", "SH21771", "SH21793", "SH21794", "SH21795", 
                   "SH21797", "SH21796", "SH21798", "SH21800", "SH21799", "SH21801")
# Define sessile rows
sessile_rows <- c("SH21706", "SH21707", "SH21708", "SH21709", "SH21710", "SH21711", 
                   "SH21712", "SH21713", "SH21714", "SH21718", "SH21719", "SH21720", 
                   "SH21721", "SH21722", "SH21723", "SH21724", "SH21725", "SH21726", 
                   "SH21700", "SH21701", "SH21702", "SH21703", "SH21704", "SH21705", 
                   "SH21730", "SH21731", "SH21732", "SH21733", "SH21734", "SH21735", 
                   "SH21715", "SH21716", "SH21717", "SH21727", "SH21728", "SH21729")

# Subset the EnMatrix_ind dataframe
EnvMatrix_motile <- EnvMatrix_ind[rownames(EnvMatrix_ind) %in% motile_rows, ]
EnvMatrix_sessile <- EnvMatrix_ind[rownames(EnvMatrix_ind) %in% sessile_rows, ]

# View the subset
head(EnvMatrix_motile)
```
# Export R objects of environmental distances 
```{r}
save(sEnvMatrix_motile, env_dist_mot, sEnvMatrix_sessile, env_dist_sess, sEnvMatrix_ind, env_dist_df, file = "./env_dist_input.Rdata")
```



Calculate motile env distances
```{r}
# Motile
sEnvMatrix_motile = scale(EnvMatrix_motile, center = T, scale = T) # Scale transform
sEnvMatrix_motile <- as.data.frame(sEnvMatrix_motile) # Convert to dataframe
env_dist_mot <- dist(sEnvMatrix_motile) # Calculate euclidean distances
boxplot(sEnvMatrix_motile) # Plot scale transformed values

#Sessile
sEnvMatrix_sessile = scale(EnvMatrix_motile, center = T, scale = T) # Scale transform
sEnvMatrix_motile <- as.data.frame(sEnvMatrix_motile) # Convert to dataframe
env_dist_mot <- dist(sEnvMatrix_motile) # Calculate euclidean distances
boxplot(sEnvMatrix_motile) # Plot scale transformed values
```

===========================================
Calculate Geographic Distances with MARMAP
===========================================
```{r}
# Load marmap
#library(marmap)
```

#Add depth to coord_sites dataframe for z dimension
# Ensure both dataframes have rownames that match exactly
rownames(coord_sites)
rownames(EnvMatrix_reduced)

# Add the depth column from EnvMatrix_reduced to coord_sites
coord_sites$depth <- EnvMatrix_reduced[rownames(coord_sites), "depth"]

# Check if the depth column was added correctly
head(coord_sites)

Calculate distances for 12 sites
```{r}
library(marmap)

# Load bathymetric data (adjust lon/lat and resolution for your region)
bathydata <- getNOAA.bathy(lon1 = -95, lon2 = -90, lat1 = 27, lat2 = 30, resolution = 0.2)

# Create a transition matrix based on bathymetric data, constrained by minimum depth
t <- trans.mat(bathydata, min.depth = -0.1)

# Load your site coordinates 
# Assuming 'coord_sites' contains the columns 'longitude' and 'latitude'
xy_sites <- as.data.frame(coord_sites[,c("longitude", "latitude")])

# Calculate the least-cost distance matrix in kilometers
leastDist.km <- lc.dist(t, xy_sites, res = "dist")
leastDist.km

# Save the distance matrix for later use
save(leastDist.km, file = 'leastDist.km.Rdata')

# Convert the 'dist' object to a matrix
leastDist.km_matrix <- as.matrix(leastDist.km)

# Convert the matrix to a data frame
leastDist.km_df <- as.data.frame(leastDist.km_matrix)

# Write the data frame to a CSV file
write.csv(leastDist.km_df, file = "./leastDist_km.csv", row.names = TRUE)

# Calculate the minimum observed distance between sites
min_dist <- min(leastDist.km)
print(min_dist)


```


Calculate geographic distance for all, motile, and arms ARMS samples in order to have the proper dimensions for down-stream Mantel tests with Bray-Curtis distances
```{r}
# Load your site coordinates for individual ARMS 
coord_sites_ind <- read.csv("./coord_sites_indv.csv",row.names=1, sep=",")

# Subset the coord_sites dataframe for motile and sessile fractions
coord_sites_mot <- coord_sites_ind[rownames(coord_sites_ind) %in% motile_rows, ]
coord_sites_sess <- coord_sites_ind[rownames(coord_sites_ind) %in% sessile_rows, ]

# Assuming 'coord_sites' contains the columns 'longitude' and 'latitude'
xy_sites_ind <- as.data.frame(coord_sites_ind[,c("longitude", "latitude")]) #all
xy_sites_mot <- as.data.frame(coord_sites_mot[,c("longitude", "latitude")]) #motile
xy_sites_sess <- as.data.frame(coord_sites_sess[,c("longitude", "latitude")]) #sessile

# Calculate the least-cost distance matrix in kilometers
leastDist.km_ind <- lc.dist(t, xy_sites_ind, res = "dist") #all
leastDist.km_mot <- lc.dist(t, xy_sites_mot, res = "dist") #motile
leastDist.km_sess <- lc.dist(t, xy_sites_sess, res = "dist") #sessile

# Save the distance matrix for later use
save(leastDist.km_ind, file = 'leastDist.km_ind.Rdata') #all
save(leastDist.km_mot, file = 'leastDist.km_mot.Rdata') #motile
save(leastDist.km_sess, file = 'leastDist.km_sess.Rdata')#sessile

# Convert the 'dist' object to a matrix
leastDist.km_ind_matrix <- as.matrix(leastDist.km_ind) #all
leastDist.km_mot_matrix <- as.matrix(leastDist.km_mot) #motile
leastDist.km_sess_matrix <- as.matrix(leastDist.km_sess) #sessile

# Convert the matrix to a data frame
leastDist.km_ind_df <- as.data.frame(leastDist.km_ind_matrix) #all
leastDist.km_mot_df <- as.data.frame(leastDist.km_mot_matrix) #motile
leastDist.km_sess_df <- as.data.frame(leastDist.km_sess_matrix) #sessile

# Write the data frame to a CSV file
write.csv(leastDist.km_ind_df, file = "./leastDist_km_ind.csv", row.names = TRUE)
write.csv(leastDist.km_mot_df, file = "./leastDist_km_motile.csv", row.names = TRUE)
write.csv(leastDist.km_sess_df, file = "./leastDist_km_sessile.csv", row.names = TRUE)

# Calculate the minimum observed distance between sites, will be zero because some sites extremely close to each other within a bank
#min_dist <- min(leastDist.km)
#print(min_dist)
```

